---
title: "Part 07 — Analyzing Visual Content: Images, Alt Text, and Link Cards"
subtitle: "Bluesky Data Science in R"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

This document is the R counterpart of the Python notebook
[Part 07 - Analyzing Visual Content](../../Notebooks/Part%2007%20-%20Analyzing%20Visual%20Content/Part%2007%20-%20Analyzing%20Visual%20Content.ipynb).
Posts are more than text: images, video, link cards, and quoted posts
arrive as typed **embeds**, and the alt text attached to images is both
an accessibility practice and a measurable research object. We collect a
panel of accounts' posts as raw lists (`clean = FALSE`), classify their
embeds, audit alt-text provision, map link-card domains, and assemble a
small documented image corpus.

### Learning objectives

* Read the anatomy of Bluesky's embed union types.
* Collect posts as raw lists that survive new, unmodeled embed types.
* Measure media composition and alt-text provision across accounts.
* Analyze link cards as an outlink map.
* Download a small image corpus responsibly, with a manifest.

## Setup

```{r libraries}
library(bskyr)
library(tidyverse)
library(jsonlite)
library(lubridate)
```

As in every part of this series, the shared helper `bsky_client.R` looks
for credentials (an `atproto.json` file or bskyr's environment
variables), authenticates if it finds them, and sets `bsky_authed`. The
`knitr::opts_chunk$set(eval = bsky_authed)` line then makes that flag the
default for the whole document: with credentials every chunk runs;
without them the document still knits, but the API chunks are skipped.

```{r connect}
source("bsky_client.R")
knitr::opts_chunk$set(eval = bsky_authed)
```

## 1. The anatomy of a Bluesky embed

A post's `record$embed` is a *union type*: exactly one of
`app.bsky.embed.images`, `...video`, `...external` (link cards),
`...record` (quote posts), or `...recordWithMedia`. The Python notebook
opens with a war story: its strictly-typed SDK *crashed* on an embed
type the platform had just shipped, and the fix was to fall back to raw
JSON. R's list-of-lists representation has that robustness built in —
`clean = FALSE` returns whatever the server sent:

```{r raw-page}
page <- bs_get_author_feed("bsky.app", limit = 30, clean = FALSE)

items <- page[[1]]$feed

map_chr(items, \(item) item$post$embed$`$type` %||% "(no embed)") |>
  table()
```

A classifier that collapses view types into analytic classes — and
shunts anything it has never seen into `"other"` instead of crashing:

```{r classify}
classify_embed <- function(post) {
  type <- post$embed$`$type` %||% return("none")
  type <- sub("#view$", "", type)
  switch(type,
    "app.bsky.embed.images" = "images",
    "app.bsky.embed.video" = "video",
    "app.bsky.embed.external" = "external",
    "app.bsky.embed.record" = "record",
    "app.bsky.embed.recordWithMedia" = "record_with_media",
    "other")
}
```

## 2. Collecting a media panel

Visual practice varies by *kind* of account, so we compare a panel:
platform, public broadcaster, local newsroom, national newsrooms,
and a policy institute. The collector below fetches each account's feed as raw pages
(`clean = FALSE` returns a list of response pages, each with a `$feed`
list of items), flattens the pages together, uses `purrr::keep()` to
retain only original posts, and extracts the `post` object from each
item:

```{r panel}
PANEL <- c("bsky.app", "npr.org", "boulderrl.bsky.social",
           "propublica.org", "nytimes.com", "knightcolumbia.org")
POSTS_PER_ACCOUNT <- 135
COLLECTED_AT <- format(now("UTC"), "%Y-%m-%dT%H:%M:%SZ")

collect_raw_posts <- function(actor, max_posts) {
  pages <- bs_get_author_feed(actor, limit = max_posts, clean = FALSE)
  items <- list_flatten(map(pages, "feed"))
  # skip reposts: the feed item carries a 'reason' explaining its presence
  originals <- keep(items, \(item) is.null(item$reason))
  map(originals, "post")
}

all_posts <- map(PANEL, \(actor) {
  posts <- collect_raw_posts(actor, POSTS_PER_ACCOUNT)
  Sys.sleep(0.1)
  posts
}) |> set_names(PANEL)

map_int(all_posts, length)
```

Now three helpers that walk a raw post. `iter_images()` returns one row
per attached image, with its alt text and declared dimensions; the
`%||%` chains handle the fact that the image list lives in different
places for plain image posts (`embed$images`) and posts that pair media
with a quote (`embed$media$images`). `external_info()` pulls the link
card and reduces its URL to a bare domain with a regular expression.
`post_row()` assembles everything into the one-row-per-post summary our
analysis runs on:

```{r media-helpers}
iter_images <- function(post) {
  view <- post$embed
  imgs <- view$images %||% view$media$images %||% list()
  map(imgs, \(img) tibble(
    alt = img$alt %||% "",
    thumb = img$thumb %||% NA_character_,
    width = img$aspectRatio$width %||% NA_integer_,
    height = img$aspectRatio$height %||% NA_integer_
  )) |> list_rbind()
}

url_domain <- function(url) {
  str_match(url, "^https?://(?:www\\.)?([^/]+)")[, 2]
}

external_info <- function(post) {
  ext <- post$embed$external
  if (is.null(ext)) return(NULL)
  tibble(title = ext$title %||% "",
         domain = url_domain(ext$uri %||% ""))
}

post_row <- function(post) {
  imgs <- iter_images(post)
  ext <- external_info(post)
  tibble(
    uri = post$uri,
    author = post$author$handle,
    embed_class = classify_embed(post),
    n_images = nrow(imgs),
    n_images_with_alt = sum(str_squish(imgs$alt) != ""),
    external_domain = if (is.null(ext)) NA_character_ else ext$domain
  )
}

df <- map(all_posts, \(posts) map(posts, post_row) |> list_rbind()) |>
  list_rbind()

count(df, embed_class, sort = TRUE)
```

## 3. How visual is the panel?

A stacked bar per account shows each one's *embed mix* — what share of
its posts are bare text, images, video, link cards, or quotes. Newsrooms
should lean on link cards; individuals on text and images:

```{r composition, fig.width=9, fig.height=4.5}
df |>
  count(author, embed_class) |>
  group_by(author) |>
  mutate(share = n / sum(n)) |>
  ggplot(aes(share, author, fill = embed_class)) +
  geom_col() +
  labs(x = "Share of posts", y = NULL, fill = "Embed type",
       title = "Embed type composition by account") +
  theme_minimal()
```

### Images per post and aspect ratios

```{r aspect, fig.width=8, fig.height=4}
images_df <- imap(all_posts, \(posts, actor) {
  map(posts, iter_images) |> list_rbind() |> mutate(author = actor)
}) |> list_rbind()

c(image_posts = sum(df$n_images > 0), images = nrow(images_df))

images_df |>
  filter(!is.na(width), !is.na(height)) |>
  mutate(aspect = width / height) |>
  ggplot(aes(aspect)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  geom_vline(xintercept = c(9 / 16, 1, 16 / 9),
             color = "crimson", linetype = "dashed") +
  labs(x = "Aspect ratio (width / height)", y = "Images",
       title = "Aspect ratios of panel images",
       subtitle = "Dashed: 9:16 portrait, 1:1 square, 16:9 landscape") +
  theme_minimal()
```

## 4. Alt text: accessibility as a measurable practice

Bluesky lets authors attach alt text to every image. Gleason et al.
(2019) found only ~0.1% of image tweets carried alt text; Bluesky's
culture is dramatically different, and we can measure it:

```{r alt-rates}
image_posts <- df |> filter(n_images > 0)

tibble(
  image_posts = nrow(image_posts),
  all_images_described = mean(image_posts$n_images_with_alt ==
                              image_posts$n_images),
  any_image_described = mean(image_posts$n_images_with_alt > 0),
  images_with_alt = mean(str_squish(images_df$alt) != "")
)
```

```{r alt-by-account, fig.width=8, fig.height=4}
images_df |>
  group_by(author) |>
  summarise(images = n(),
            alt_rate = mean(str_squish(alt) != "")) |>
  ggplot(aes(alt_rate, fct_reorder(author, alt_rate))) +
  geom_col(fill = "seagreen") +
  geom_text(aes(label = scales::percent(alt_rate, 1)), hjust = -0.1) +
  geom_vline(xintercept = 0.001, color = "crimson", linetype = "dashed") +
  scale_x_continuous(limits = c(0, 1.15)) +
  labs(x = "Share of posted images with alt text", y = NULL,
       title = "Alt-text provision rate by account",
       subtitle = "Dashed red line: Twitter 2019 baseline (0.1%)") +
  theme_minimal()
```

### How much do people write?

```{r alt-length, fig.width=8, fig.height=4}
alt_lengths <- images_df |>
  filter(str_squish(alt) != "") |>
  mutate(alt_len = nchar(alt))

ggplot(alt_lengths, aes(alt_len)) +
  geom_histogram(bins = 30, fill = "seagreen", color = "white") +
  geom_vline(xintercept = median(alt_lengths$alt_len),
             linetype = "dashed") +
  labs(x = "Alt text length (characters)", y = "Images",
       title = paste("Length of", nrow(alt_lengths), "non-empty alt texts")) +
  theme_minimal()

summary(alt_lengths$alt_len)
```

## 5. Link cards: who points where?

External embeds carry a URL, title, and description — an outlink map of
each account's information diet (and self-promotion):

```{r domains}
external_posts <- df |> filter(!is.na(external_domain))

external_posts |>
  count(external_domain, sort = TRUE) |>
  mutate(share = round(n / nrow(external_posts), 3)) |>
  slice_head(n = 10)
```

```{r self-links}
OWN_DOMAINS <- list(
  "bsky.app" = c("bsky.app", "bsky.social"),
  "npr.org" = "npr.org",
  "boulderrl.bsky.social" = "boulderreportinglab.org",
  "propublica.org" = "propublica.org",
  "nytimes.com" = c("nytimes.com", "nyti.ms"),
  "knightcolumbia.org" = "knightcolumbia.org"
)

external_posts |>
  group_by(author) |>
  summarise(own_domain_share = mean(map2_lgl(
    external_domain, author,
    \(domain, a) any(endsWith(domain, OWN_DOMAINS[[a]]))
  )))
```

## 6. Building a small image corpus

Visual content analysis needs the images themselves. Download a *small*,
documented sample of thumbnails (never originals at scale), with a
manifest tying every file to its post, author, and alt text. `walk2()`
is `map2()`'s sibling for side effects — it loops over URL/destination
pairs calling `download.file()`, sleeping half a second between
downloads out of politeness to the CDN:

```{r download}
dir.create("data/images", recursive = TRUE, showWarnings = FALSE)

manifest <- imap(all_posts, \(posts, actor) {
  map(posts, \(post) {
    imgs <- iter_images(post)
    if (nrow(imgs) == 0) return(NULL)
    mutate(imgs[1, ], author = actor, uri = post$uri)
  }) |> list_rbind()
}) |>
  list_rbind() |>
  filter(!is.na(thumb)) |>
  group_by(author) |>
  slice_head(n = 2) |>
  ungroup() |>
  mutate(file = file.path("data/images",
                          paste0("img_", row_number(), ".jpg")))

walk2(manifest$thumb, manifest$file, \(url, dest) {
  try(download.file(url, dest, mode = "wb", quiet = TRUE), silent = TRUE)
  Sys.sleep(0.5)
})

write_csv(select(manifest, -thumb), "data/images_manifest.csv")

c(images_saved = sum(file.exists(manifest$file)),
  total_kb = round(sum(file.size(manifest$file), na.rm = TRUE) / 1024))
```

## 7. From pixels to measurements: CLIP embeddings

Modern visual content analysis encodes images into the same vector
space as text (CLIP), enabling zero-shot labeling and clustering. From
R, CLIP is reachable through [reticulate](https://rstudio.github.io/reticulate/)
and the `sentence-transformers` Python package; the Python notebook
runs this exact workflow natively. The design point stands regardless of
language: define a visual codebook in words, embed both, and label by
cosine similarity.

```{r clip, eval=FALSE}
library(reticulate)
st <- import("sentence_transformers")
model <- st$SentenceTransformer("clip-ViT-B-32")
image_embeddings <- model$encode(manifest$file)
label_embeddings <- model$encode(c(
  "a screenshot of text or a user interface",
  "a photograph of one or more people",
  "a chart, map, or data visualization"))
```

## 8. Saving and reloading the archive

```{r save}
write_csv(df, "data/media_summary.csv")
write_json(
  list(collected_at = COLLECTED_AT, panel = PANEL,
       posts_per_account = POSTS_PER_ACCOUNT, reposts_excluded = TRUE,
       endpoint = "app.bsky.feed.getAuthorFeed (raw lists)"),
  "data/collection_metadata.json", pretty = TRUE, auto_unbox = TRUE)

reloaded <- read_csv("data/media_summary.csv", show_col_types = FALSE)
count(reloaded, embed_class, sort = TRUE)
```

## Ethics and limitations

Images often contain *faces* — analyzing, clustering, or republishing
them implicates people far beyond the account that posted. Download
sparingly, prefer thumbnails, keep manifests so corpora are auditable
and deletable, and never build face-recognition pipelines on collected
social images. Alt text is written *for* blind and low-vision users;
treating it as research data should not crowd out that purpose. Our
panel is six hand-picked accounts — illustrative, not representative —
and embed metadata (aspect ratios, alt text) is only as accurate as the
posting client made it.

## Exercises

1. Add two accounts you follow to the panel and recompute the alt-text
   table. Where do they fall?
2. Classify alt texts as substantive vs. perfunctory (e.g., length
   thresholds plus keyword rules) and re-estimate "real" provision.
3. Map quoted-post authors from `record` embeds: whom does each panel
   account amplify by quoting?
4. Compare engagement (likes) of image posts vs. text-only posts within
   each account.
5. With the magick package, compute mean brightness for the downloaded
   thumbnails and compare across accounts.

## Research project ideas

* **The accessibility gap.** Measure alt-text provision across account
  types (news, government, individual) at scale and model its predictors
  — following Gleason et al. (2019) on an open platform.
* **Visual framing of news.** For newsrooms on the panel, classify
  link-card thumbnails (people vs. charts vs. scenes) and compare visual
  framing of the same story across outlets.

## References

### Scholarship

* Gleason, C., Carrington, P., Cassidy, C., Morris, M. R., Kitani, K. M., & Bigham, J. P. (2019). "It's almost like they're trying to hide it": How user-provided image descriptions have failed to make Twitter accessible. *WWW '19*.
* Casas, A., & Williams, N. W. (2019). Images that matter: Online protests and the mobilizing role of pictures. *Political Research Quarterly*, 72(2), 360–375.
* Radford, A., et al. (2021). Learning transferable visual models from natural language supervision (CLIP). *ICML*.

### Documentation

* Embed lexicons: <https://docs.bsky.app/docs/advanced-guides/posts#embeds>
* reticulate: <https://rstudio.github.io/reticulate/>
* bskyr: <https://christophertkenny.com/bskyr/>
