---
title: "Part 05 — Time Series and Causal Inference: Event Studies on Bluesky"
subtitle: "Bluesky Data Science in R"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

This document is the R counterpart of the Python notebook
[Part 05 - Time Series and Causal Inference](../../Notebooks/Part%2005%20-%20Time%20Series%20and%20Causal%20Inference/Part%2005%20-%20Time%20Series%20and%20Causal%20Inference.ipynb).
It turns complete posting histories into time series, describes their
rhythms, and then asks a harder question: did a real-world event *cause*
a change in behavior? The same two quasi-experimental designs —
interrupted time series and difference-in-differences — are fit here
with base R's `lm()` in place of statsmodels.

### Learning objectives

* Collect complete posting histories and handle timestamps and timezones.
* Describe diurnal, weekly, and seasonal posting rhythms.
* Fit and interpret an interrupted time series (ITS) regression.
* Fit difference-in-differences (DiD) and check the parallel-trends assumption.
* Articulate the threats to validity that observational designs face.

## Setup

```{r libraries}
library(bskyr)
library(tidyverse)
library(jsonlite)
library(lubridate)
```

As in every part of this series, the shared helper `bsky_client.R` looks
for credentials (an `atproto.json` file or bskyr's environment
variables), authenticates if it finds them, and sets `bsky_authed`. The
`knitr::opts_chunk$set(eval = bsky_authed)` line then makes that flag the
default for the whole document: with credentials every chunk runs;
without them the document still knits, but the API chunks are skipped.

```{r connect}
source("bsky_client.R")
knitr::opts_chunk$set(eval = bsky_authed)
```

## 1. Collecting complete posting histories

The event we will study is the November 2024 US election and the surge
of Twitter-to-Bluesky migration that followed it. We need two complete
histories: a *treated* account (The New York Times, a newsroom whose
audience and posting we suspect changed when millions of users arrived)
and a *comparison* account:

```{r collect}
TREATED <- "nytimes.com"
COMPARISON <- "bsky.app"

treated_feed <- bs_get_author_feed(TREATED, limit = 40000)
comparison_feed <- bs_get_author_feed(COMPARISON, limit = 10000)

c(treated = nrow(treated_feed), comparison = nrow(comparison_feed))
```

These histories include everything the feed returns — original posts,
replies, and reposts are all *actions* in the account's timeline, which
is what we want for an activity study.

## 2. Timestamps and timezones

API timestamps are UTC; humans live in local time, and diurnal rhythms
only make sense in the timezone where the humans behind the account
work — for a New York newsroom, Eastern time. The helper below
does three jobs: it classifies each feed item as a post, reply, or
repost (`case_when()` is dplyr's vectorized if/else ladder); it parses
the ISO timestamp with `ymd_hms()` and converts it to local time with
`with_tz()`; and it returns one tidy row per action, oldest first:

```{r timestamps}
history_frame <- function(feed, actor, tz = "America/New_York") {
  is_repost <- if ("reason_type" %in% names(feed))
    !is.na(feed$reason_type) else rep(FALSE, nrow(feed))
  is_reply <- if ("reply_parent_uri" %in% names(feed))
    !is.na(feed$reply_parent_uri) else rep(FALSE, nrow(feed))
  feed |>
    mutate(
      actor = actor,
      action = case_when(is_repost ~ "repost",
                         is_reply ~ "reply",
                         .default = "post"),
      created_at = ymd_hms(
        coalesce(map_chr(record, "created_at", .default = NA), indexed_at),
        quiet = TRUE) |> with_tz(tz)
    ) |>
    filter(!is.na(created_at)) |>
    select(actor, uri, action, created_at) |>
    arrange(created_at)
}

treated_posts <- history_frame(treated_feed, TREATED)
comparison_posts <- history_frame(comparison_feed, COMPARISON)

count(treated_posts, action)
```

## 3. Descriptive time series: the rhythms of posting

### Daily counts and a rolling mean

Counting actions per day gives a noisy series; a *rolling mean* smooths
it so the trend is visible. Two details: `complete()` fills in
zero-count days (a day with no posts is data, not a gap!), and
`stats::filter()` with equal weights of 1/k computes a centered k-day
moving average — base R's one-line rolling mean:

```{r daily, fig.width=10, fig.height=4}
daily_counts <- function(posts) {
  posts |>
    count(date = as_date(created_at)) |>
    complete(date = seq(min(date), max(date), by = "day"),
             fill = list(n = 0))
}

rolling_mean <- function(x, k = 28) {
  as.numeric(stats::filter(x, rep(1 / k, k), sides = 2))
}

treated_daily <- daily_counts(treated_posts)

treated_daily |>
  mutate(rolling = rolling_mean(n)) |>
  ggplot(aes(date)) +
  geom_line(aes(y = n), color = "lightsteelblue", linewidth = 0.4) +
  geom_line(aes(y = rolling), color = "navy", linewidth = 1) +
  labs(x = "Date", y = "Actions per day",
       title = paste("Daily activity:", TREATED),
       subtitle = "28-day centered rolling mean in navy") +
  theme_minimal()
```

### Weekly counts

```{r weekly, fig.width=10, fig.height=4}
weekly_counts <- function(posts) {
  posts |>
    count(week = floor_date(created_at, "week")) |>
    complete(week = seq(min(week), max(week), by = "week"),
             fill = list(n = 0))
}

treated_weekly <- weekly_counts(treated_posts)

ggplot(treated_weekly, aes(week, n)) +
  geom_line(color = "navy") +
  labs(x = "Week", y = "Actions per week",
       title = paste("Weekly activity:", TREATED)) +
  theme_minimal()
```

### Hour-of-day and day-of-week profiles

```{r rhythms, fig.width=10, fig.height=4}
treated_posts |>
  count(hour = hour(created_at)) |>
  ggplot(aes(hour, n)) +
  geom_line(color = "navy") + geom_point(color = "navy") +
  scale_x_continuous(breaks = seq(0, 22, 2)) +
  labs(x = "Hour of day (America/New_York)", y = "Total actions",
       title = paste("Diurnal profile:", TREATED)) +
  theme_minimal()

treated_posts |>
  count(dow = wday(created_at, label = TRUE, week_start = 1)) |>
  ggplot(aes(dow, n)) +
  geom_col(fill = "navy", alpha = 0.85) +
  labs(x = "Day of week", y = "Total actions",
       title = paste("Weekly rhythm:", TREATED)) +
  theme_minimal()
```

### Year-over-year seasonality

```{r yoy, fig.width=10, fig.height=4}
treated_posts |>
  count(year = factor(year(created_at)), month = month(created_at)) |>
  ggplot(aes(month, n, color = year)) +
  geom_line() + geom_point() +
  scale_x_continuous(breaks = 1:12) +
  labs(x = "Month", y = "Actions per month", color = "Year",
       title = paste("Year-over-year monthly activity:", TREATED)) +
  theme_minimal()
```

## 4. An event: the November 2024 Bluesky surge

```{r event-line, fig.width=10, fig.height=4}
EVENT <- as_date("2024-11-05")

ggplot(treated_weekly, aes(week, n)) +
  geom_line(color = "navy") +
  geom_vline(xintercept = EVENT, color = "crimson", linetype = "dashed") +
  labs(x = "Week", y = "Actions per week",
       title = paste("Weekly activity with event line:", TREATED)) +
  theme_minimal()
```

A before/after eyeball is not a causal estimate: activity has trends and
seasonality, and anything else that changed in November 2024 is bundled
into the comparison. The designs below discipline that comparison.

## 5. Interrupted time series

ITS fits the segmented regression
*y* = β₀ + β₁·t + β₂·post + β₃·(t − t_event)·post + ε:
a pre-period trend, a level shift at the event, and a trend change after
it. The counterfactual is the pre-trend extended forward.

Read the chunk below as building the regression's design matrix by hand:
`t` numbers the weeks 1, 2, 3, ...; `post` is 0 before the event and 1
after; and `time_since` counts weeks elapsed since the event (0 before
it). Then `lm()` — R's workhorse linear-regression function — fits the
model, and `summary()` prints coefficients, standard errors, and
p-values:

```{r its}
its_data <- treated_weekly |>
  mutate(
    t = row_number(),
    post = as.integer(as_date(week) >= EVENT),
    time_since = pmax(0, t - min(t[post == 1]) + 1) * post
  )

its_fit <- lm(n ~ t + post + time_since, data = its_data)
summary(its_fit)
```

```{r its-table}
its_table <- broom::tidy(its_fit, conf.int = TRUE) |>
  mutate(term = c("intercept", "trend", "level_shift", "trend_change"))
its_table
```

```{r its-plot, fig.width=10, fig.height=4}
its_data |>
  mutate(
    fitted = fitted(its_fit),
    counterfactual = coef(its_fit)["(Intercept)"] + coef(its_fit)["t"] * t
  ) |>
  ggplot(aes(week)) +
  geom_line(aes(y = n), color = "lightsteelblue", linewidth = 0.5) +
  geom_line(aes(y = fitted), color = "navy", linewidth = 1) +
  geom_line(aes(y = counterfactual), color = "gray40",
            linetype = "dashed") +
  geom_vline(xintercept = EVENT, color = "crimson", linetype = "dashed") +
  labs(x = "Week", y = "Actions per week",
       title = paste("Interrupted time series:", TREATED),
       subtitle = "Dashed gray: pre-trend extended as counterfactual") +
  theme_minimal()
```

### An honest aside: autocorrelation

Weekly counts are autocorrelated — this week predicts next week — which
makes OLS standard errors too small. Check with Durbin–Watson and refit
with heteroskedasticity-and-autocorrelation-consistent (HAC) errors:

```{r hac}
resids <- residuals(its_fit)
dw <- sum(diff(resids)^2) / sum(resids^2)
round(dw, 3)  # 2 = no autocorrelation; < 2 = positive autocorrelation

if (requireNamespace("sandwich", quietly = TRUE) &&
    requireNamespace("lmtest", quietly = TRUE)) {
  lmtest::coeftest(its_fit, vcov = sandwich::NeweyWest(its_fit, lag = 4))
} else {
  message("Install the sandwich and lmtest packages for HAC standard errors.")
}
```

## 6. Difference-in-differences

ITS extrapolates the treated unit's own trend; DiD instead uses a
*comparison unit* to absorb shocks common to both. Its identifying
assumption is **parallel trends**: absent the event, the two series
would have moved together. Always plot the pre-trends first:

```{r pretrends, fig.width=10, fig.height=4}
comparison_weekly <- weekly_counts(comparison_posts)

bind_rows(
  mutate(treated_weekly, account = TREATED),
  mutate(comparison_weekly, account = COMPARISON)
) |>
  ggplot(aes(week, n, color = account)) +
  geom_line() +
  geom_vline(xintercept = EVENT, color = "crimson", linetype = "dashed") +
  labs(x = "Week", y = "Actions per week",
       title = "Pre-trend plot: treated vs. comparison account") +
  theme_minimal()
```

The DiD estimate itself can be computed two equivalent ways, and seeing
both demystifies the regression: first as arithmetic on the 2×2 table of
means (treated/comparison × pre/post), then as the interaction
coefficient in `lm(n ~ treated * post)`, which adds standard errors:

```{r did}
panel <- bind_rows(
  mutate(treated_weekly, treated = 1L),
  mutate(comparison_weekly, treated = 0L)
) |>
  mutate(post = as.integer(as_date(week) >= EVENT))

# The 2x2 of means, and the DiD by hand
panel |>
  group_by(treated, post) |>
  summarise(mean_actions = mean(n), .groups = "drop") |>
  pivot_wider(names_from = post, values_from = mean_actions,
              names_prefix = "post_")

did_fit <- lm(n ~ treated * post, data = panel)
broom::tidy(did_fit, conf.int = TRUE) |>
  mutate(term = c("intercept", "treated", "post", "treated_post"))
```

The `treated_post` interaction is the DiD estimate.

### Why this DiD should not be believed

Our comparison unit is the platform's *official account* — its posting
is driven by product launches, not by the news cycle that drives a
newsroom's account, and the November surge changed *its* behavior too.
A comparison that responds to the treatment violates the design. The
lesson generalizes: DiD is only as good as the comparison group.

## 7. Building a better comparison group: matching on profile covariates

A principled comparison pool: accounts similar to the treated unit on
observables. Here is nearest-neighbor matching on (logged) followers,
posts, and account age. We log the heavy-tailed counts so one
mega-account doesn't dominate, standardize each feature with `scale()`
(subtract the mean, divide by the standard deviation — so all three are
in comparable units), and rank candidates by Euclidean distance to the
treated account:

```{r matching}
candidates <- bs_get_follows(TREATED, limit = 50)

pool <- bs_get_profile(c(TREATED, head(candidates$handle, 24))) |>
  mutate(account_age_days = as.numeric(
    difftime(now("UTC"), ymd_hms(created_at, quiet = TRUE), units = "days"))) |>
  filter(!is.na(followers_count), !is.na(account_age_days))

features <- pool |>
  transmute(handle,
            log_followers = log1p(followers_count),
            log_posts = log1p(posts_count),
            account_age_days) |>
  mutate(across(-handle, scale))

target <- features |> filter(handle == TREATED)

features |>
  filter(handle != TREATED) |>
  mutate(distance = sqrt((log_followers - target$log_followers)^2 +
                         (log_posts - target$log_posts)^2 +
                         (account_age_days - target$account_age_days)^2)) |>
  left_join(select(pool, handle, followers_count, posts_count),
            by = "handle") |>
  slice_min(distance, n = 5) |>
  select(handle, followers_count, posts_count, distance)
```

A full design would collect histories for several matched accounts and
average them — see the MatchIt package for the industrial version.

## 8. Keyword volume around an event

`bs_search_posts()` with `since`/`until` windows builds a topic-level
series — e.g., weekly volume of posts mentioning "twitter" around the
election. Capped to stay polite:

```{r keyword-volume, fig.width=10, fig.height=4}
edges <- seq(as_date("2024-10-01"), as_date("2024-12-31"), by = "week")

volume <- map2_int(head(edges, -1), tail(edges, -1), \(from, to) {
  hits <- tryCatch(
    bs_search_posts("twitter",
                    since = format(from, "%Y-%m-%dT00:00:00Z"),
                    until = format(to, "%Y-%m-%dT00:00:00Z"),
                    limit = 1000),
    error = function(e) NULL)
  Sys.sleep(0.2)
  if (is.null(hits)) NA_integer_ else nrow(hits)
})

tibble(week = tail(edges, -1), posts = volume) |>
  ggplot(aes(week, posts)) +
  geom_line(color = "navy") +
  geom_vline(xintercept = as_date("2024-11-05"),
             color = "crimson", linetype = "dashed") +
  labs(y = "Posts per week (capped at 1,000)",
       title = "Weekly volume of posts matching 'twitter'") +
  theme_minimal()
```

## 9. Saving and reloading the data

```{r save}
dir.create("data", showWarnings = FALSE)

daily_both <- bind_rows(
  mutate(daily_counts(treated_posts), account = TREATED),
  mutate(daily_counts(comparison_posts), account = COMPARISON)
)

write_csv(daily_both, "data/daily_counts.csv")
write_csv(its_table, "data/its_results.csv")
write_json(
  list(collected_at = format(now("UTC"), "%Y-%m-%dT%H:%M:%SZ"),
       accounts = list(treated = TREATED, comparison = COMPARISON),
       event = "2024-11-05", timezone = "America/New_York",
       endpoint = "app.bsky.feed.getAuthorFeed"),
  "data/collection_metadata.json", pretty = TRUE, auto_unbox = TRUE)

reloaded <- read_csv("data/daily_counts.csv", show_col_types = FALSE)
nrow(reloaded) == nrow(daily_both)
```

## 10. Research design: what it takes to claim causation

Neither design above randomizes anything. ITS assumes nothing else
changed at the event date that affects the outcome; DiD assumes parallel
trends. Both fail silently when violated. The discipline is: specify the
design *before* looking at the post-period; plot pre-trends; run placebo
tests (fake event dates, unaffected outcomes); and treat estimates as
"the effect under these assumptions," stated plainly. Natural
experiments on platforms — outages, feature launches, deplatformings —
are the cleanest events because their timing is plausibly exogenous to
any one account's behavior.

## Ethics and limitations

Activity histories reveal sleep, work, travel, and life disruptions —
treat even public timing data as sensitive, aggregate when reporting,
and avoid building behavioral profiles of private individuals. Our data
are two accounts chosen for pedagogy, weekly bins hide within-week
dynamics, the author feed omits deleted posts (survivorship in time!),
and every caveat from Sections 5–7 applies to any causal language.

## Exercises

1. Refit the ITS with daily instead of weekly counts. How do the
   estimates and their standard errors change?
2. Run a placebo ITS with a fake event date of 2024-06-01. Is the
   level shift "significant"? What does that teach you?
3. Restrict the panel to ±12 weeks around the event and refit the DiD.
4. Use the matching table to pick the nearest neighbor, collect its
   history, and refit the DiD with it as the comparison.
5. Decompose the treated series by `action` type — did replies and
   reposts respond to the event differently?

## Research project ideas

* **Migration waves.** Using account creation dates from profiles,
  identify accounts that joined during the November 2024 surge and
  compare their retention curves to earlier cohorts.
* **Event studies of platform shocks.** Build an ITS around a Bluesky
  feature launch (e.g., DMs or video) using keyword volume as the
  outcome, with placebo dates as robustness checks.

## References

### Scholarship

* Bernal, J. L., Cummins, S., & Gasparrini, A. (2017). Interrupted time series regression for the evaluation of public health interventions. *International Journal of Epidemiology*, 46(1), 348–355.
* Angrist, J. D., & Pischke, J.-S. (2009). *Mostly Harmless Econometrics*. Princeton University Press.
* Golder, S. A., & Macy, M. W. (2011). Diurnal and seasonal mood vary with work, sleep, and daylength across diverse cultures. *Science*, 333(6051), 1878–1881.
* Cunningham, S. (2021). *Causal Inference: The Mixtape*. Yale University Press. <https://mixtape.scunning.com/>

### Documentation

* lubridate: <https://lubridate.tidyverse.org/>
* sandwich (HAC errors): <https://sandwich.R-Forge.R-project.org/>
* MatchIt: <https://kosukeimai.github.io/MatchIt/>
* bskyr: <https://christophertkenny.com/bskyr/>
