---
title: "Part 10 — Real-Time Streams and Bots: Drinking from the Firehose"
subtitle: "Bluesky Data Science in R"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

This document is the R counterpart of the Python notebook
[Part 10 - Real-Time Streams and Bots](../../Notebooks/Part%2010%20-%20Real-Time%20Streams%20and%20Bots/Part%2010%20-%20Real-Time%20Streams%20and%20Bots.ipynb).
Every previous part asked the network about its past; this one watches
it happen. We subscribe to the **Jetstream** — Bluesky's friendly JSON
event stream — collect a live sample, characterize a minute of the whole
network, and then close the loop: a minimal, well-behaved bot that reads
its notifications and (only in dry-run) replies.

### Learning objectives

* Distinguish retrospective collection from streaming collection.
* Subscribe to the Jetstream with server-side collection filters.
* Practice data minimization on high-volume streams.
* Understand the raw firehose and why Jetstream usually suffices.
* Design a bot with rate limits, disclosure, and a kill switch.

## Setup

Streaming needs a websocket client; bot actions need bskyr. The
[websocket](https://cran.r-project.org/package=websocket) package is the
one extra dependency of this part:

```{r libraries}
library(bskyr)
library(tidyverse)
library(jsonlite)
library(lubridate)

has_websocket <- requireNamespace("websocket", quietly = TRUE)
```

As in every part, the shared helper `bsky_client.R` looks for
credentials and sets `bsky_authed`. This part's gating differs from the
others in one interesting way: **the Jetstream requires no
authentication at all** — anyone can listen to the public event stream —
so the streaming chunks are gated on having the websocket package rather
than on credentials. Only the bot sections at the end, which read
notifications and (hypothetically) post, need an account:

```{r connect}
source("bsky_client.R")
# Jetstream needs no authentication; bot sections do.
knitr::opts_chunk$set(eval = has_websocket)
```

## 1. The Jetstream: the network's pulse as friendly JSON

The raw firehose (`com.atproto.sync.subscribeRepos`) speaks CBOR and CAR
archives; **Jetstream** is a public service that re-serves it as plain
JSON over a websocket, with server-side filters so you receive only the
record types you ask for:

```{r jetstream-url}
JETSTREAM_URL <- "wss://jetstream2.us-east.bsky.network/subscribe"

WANTED_COLLECTIONS <- c(
  "app.bsky.feed.post",     # posts (and replies and quotes)
  "app.bsky.feed.like",     # likes
  "app.bsky.graph.follow"   # follows
)

build_jetstream_url <- function(collections) {
  paste0(JETSTREAM_URL, "?",
         paste0("wantedCollections=", collections, collapse = "&"))
}

build_jetstream_url(WANTED_COLLECTIONS)
```

### Anatomy of a Jetstream event

Each event is one JSON object: the actor's `did`, a microsecond
`time_us`, and for commits the `collection`, `operation`
(create/update/delete), and the full `record`. No decoding required —
that is the entire pitch.

## 2. Collecting a live sample

A bounded collector: listen for a fixed duration with a hard event cap —
*every* streaming collector needs both. Websockets are event-driven, so
the code reads differently from everything else in this series: we
register an `onMessage` callback that fires once per arriving event, and
then run a small loop that hands control to the event system
(`later::run_now()`) until the deadline. The `<<-` assignments let the
callback update the shared `events` list that lives outside it:

```{r collect}
collect_events <- function(duration_seconds = 60, max_events = 20000,
                           collections = WANTED_COLLECTIONS) {
  events <- vector("list", max_events)
  n <- 0
  done <- FALSE
  ws <- websocket::WebSocket$new(build_jetstream_url(collections),
                                 autoConnect = FALSE)
  ws$onMessage(function(event) {
    if (n >= max_events) {
      done <<- TRUE
      return()
    }
    n <<- n + 1
    events[[n]] <<- fromJSON(event$data, simplifyVector = FALSE)
  })
  ws$onClose(function(event) done <<- TRUE)
  ws$connect()
  deadline <- Sys.time() + duration_seconds
  while (!done && Sys.time() < deadline) {
    later::run_now(0.25)
  }
  ws$close()
  later::run_now(0.5)
  compact(events[seq_len(n)])
}

collection_started <- now("UTC")
events <- collect_events(duration_seconds = 60)

length(events)
```

### How fast is the network moving?

```{r rate}
times_us <- map_dbl(events, "time_us")
window_seconds <- (max(times_us) - min(times_us)) / 1e6

c(events = length(events),
  seconds = round(window_seconds, 1),
  events_per_second = round(length(events) / window_seconds, 1))
```

## 3. What a minute of Bluesky looks like

Flatten each event to five fields. Note what we keep: **hashed** DIDs
and *no post text* — Section 5 explains why. The hash (SHA-256 via the
digest package, truncated) is one-way: it lets us count how many events
came from the same actor without being able to say *which* actor, which
is exactly the property data minimization wants:

```{r flatten}
hash_did <- function(did) {
  map_chr(did, \(d) substr(digest::digest(d, algo = "sha256"), 1, 16))
}

events_df <- map(events, \(e) tibble(
  author_hash = hash_did(e$did),
  timestamp = as_datetime(e$time_us / 1e6),
  kind = e$kind %||% NA_character_,
  operation = e$commit$operation %||% NA_character_,
  collection = e$commit$collection %||% NA_character_
)) |> list_rbind()

head(events_df)
```

### The event mix

```{r mix}
events_df |>
  count(collection, sort = TRUE) |>
  mutate(share = round(n / sum(n), 3))

count(events_df, operation, sort = TRUE)
```

Likes outnumber posts severalfold — most platform activity is reaction,
not production. Deletes flow past constantly too, which is why archives
and live streams never quite agree.

### Events per second over time

```{r eps, fig.width=10, fig.height=4}
per_second <- events_df |>
  mutate(second = floor_date(timestamp, "second")) |>
  count(second, collection) |>
  # first and last seconds are partial windows
  filter(second > min(second), second < max(second))

ggplot(per_second, aes(second, n, color = collection)) +
  geom_line() +
  labs(x = "Time (UTC)", y = "Events per second",
       title = paste("Live Bluesky events,",
                     format(collection_started, "%Y-%m-%d %H:%M UTC"))) +
  theme_minimal()
```

### Post-level features (without the posts)

```{r post-features}
posts_live <- map(events, \(e) {
  commit <- e$commit
  if (is.null(commit) || commit$operation != "create" ||
      commit$collection != "app.bsky.feed.post") {
    return(NULL)
  }
  record <- commit$record
  tibble(
    author_hash = hash_did(e$did),
    timestamp = as_datetime(e$time_us / 1e6),
    lang = unlist(record$langs)[1] %||% NA_character_,
    is_reply = !is.null(record$reply),
    has_embed = !is.null(record$embed),
    has_image = identical(record$embed$`$type`, "app.bsky.embed.images")
  )
}) |> list_rbind()

posts_live |>
  summarise(across(c(is_reply, has_embed, has_image), mean))
```

```{r langs, fig.width=8, fig.height=4}
posts_live |>
  mutate(lang = replace_na(lang, "(none declared)")) |>
  count(lang, sort = TRUE) |>
  slice_head(n = 10) |>
  ggplot(aes(n, fct_reorder(lang, n))) +
  geom_col(fill = "steelblue") +
  labs(x = "New posts in window", y = NULL,
       title = "Declared languages of live posts (top 10)") +
  theme_minimal()
```

## 4. From one minute to the whole network

```{r extrapolate}
events_df |>
  count(collection) |>
  mutate(per_second = n / window_seconds,
         per_hour = round(per_second * 3600),
         per_day = round(per_second * 86400))
```

### What a 1% sample would mean

Twitter's "Decahose" and 1% "Spritzer" defined a decade of research. A
1% sample of this window keeps only a few posts — enough for trends,
hopeless for rare events. Sample-size arithmetic *before* collection is
the difference between a design and a wish.

## 5. Data minimization in practice

A naive stream collector is a surveillance machine: full text, all
identities, at network scale. We keep aggregate per-second counts and an
anonymized post sample — enough to reproduce every figure above, nothing
that profiles an individual:

```{r minimize}
dir.create("data", showWarnings = FALSE)

per_second |>
  pivot_wider(names_from = collection, values_from = n, values_fill = 0) |>
  write_csv("data/stream_summary.csv")

set.seed(42)
posts_live |>
  slice_sample(n = min(200, nrow(posts_live))) |>
  arrange(timestamp) |>
  write_csv("data/stream_sample.csv")

fs::dir_info("data")[, c("path", "size")]
```

## 6. The raw firehose: `com.atproto.sync.subscribeRepos`

The full firehose carries every repository commit as CBOR-encoded CAR
blocks — cryptographically verifiable, and laborious to decode. R has no
mature decoder; the Python notebook demonstrates one in ~10 lines with
the `atproto` SDK. The practical guidance is language-independent: use
Jetstream unless you need verifiability or collections Jetstream won't
filter, and budget for reconnection logic in anything long-running.

## 7. Building a bot: closing the loop

A bot is a loop: read notifications, decide, (maybe) write. The decision
logic should be pure — testable with no network at all:

```{r bot-logic, eval=TRUE}
build_reply <- function(text) {
  if (str_detect(str_to_lower(text), "\\bping\\b")) "pong 🏓" else NA_character_
}

c(build_reply("hey @examplebot.bsky.social ping"),
  build_reply("lovely weather today"))   # NA: silence is a feature
```

### Reading the bot's inbox

Notifications are private to the account, so this — like everything
below — requires authentication:

```{r notifications, eval=FALSE}
notifications <- bs_get_notifications(limit = 25)

count(notifications, reason, sort = TRUE)
mentions <- filter(notifications, reason == "mention")
```

### One full cycle, as a dry run

The cycle enforces every safety invariant: a cap on replies per run, a
pause between sends, persistent deduplication so restarts never
double-reply, and a `dry_run` flag that defaults to TRUE. **This chunk
is never executed when knitting** (`eval=FALSE`) — write actions are
deliberate acts:

```{r bot-cycle, eval=FALSE}
MAX_REPLIES_PER_RUN <- 3
SECONDS_BETWEEN_REPLIES <- 5
ANSWERED_FILE <- "data/answered_notifications.json"

run_bot_once <- function(dry_run = TRUE) {
  answered <- if (file.exists(ANSWERED_FILE))
    unlist(read_json(ANSWERED_FILE)) else character(0)

  mentions <- bs_get_notifications(limit = 25) |>
    filter(reason == "mention", !uri %in% answered) |>
    mutate(text = map_chr(record, "text", .default = "")) |>
    mutate(reply = map_chr(text, build_reply)) |>
    filter(!is.na(reply)) |>
    slice_head(n = MAX_REPLIES_PER_RUN)

  for (i in seq_len(nrow(mentions))) {
    if (dry_run) {
      message("DRY RUN - would reply to ", mentions$uri[[i]],
              ": ", mentions$reply[[i]])
    } else {
      bs_post(text = mentions$reply[[i]],
              reply = bs_uri_to_url(mentions$uri[[i]]))
      Sys.sleep(SECONDS_BETWEEN_REPLIES)
    }
    answered <- c(answered, mentions$uri[[i]])
  }
  write_json(as.list(answered), ANSWERED_FILE)
}

run_bot_once(dry_run = TRUE)
```

### Scheduling the bot

Production bots are scheduled scripts, not notebooks: `Rscript bot.R`
from cron (or the [cronR](https://cran.r-project.org/package=cronR)
package), every few minutes, with logging.

### Write rate limits: the points system

Bluesky meters writes in points (creates cost 3, deletes 1) with hourly
and daily budgets per account. A polite bot stays far inside them; the
caps above are not optional decoration.

### What your bot must never do

Reply to strangers uninvited at scale, impersonate a human, follow/like
en masse, or run from your personal account. Bluesky norms expect bots
to be **disclosed as bots in their profile**, opt-in by mention, and
removable: a kill switch (stop replying when anyone says "stop") is part
of the minimum viable bot.

## 8. Research designs with streams

Streams enable designs retrospection cannot: measuring deletion rates
(what fraction of posts vanish within an hour?), event detection as news
breaks, true prospective sampling for cascade studies (Part 04's
selection problem dissolves when you sample posts *at birth*), and
platform-level activity baselines for event studies (Part 05). Each
inherits the stream's blind spot: you see only what happens while you
listen.

## Ethics and limitations

Streaming magnifies every issue from Part 01: you observe *everyone*,
including people who will delete what they wrote minutes later. Minimize
ruthlessly (we stored no text and no raw identities), aggregate before
sharing, and treat deletion as a signal to honor, not a nuisance. Bots
act in public: disclosure, rate limits, and opt-in interaction are
ethical requirements, not technical suggestions. And our one-minute
window is a vignette, not an estimate — time-of-day and day-of-week
effects swamp short samples.

## Exercises

1. Re-run the collector at a different time of day and compare
   events-per-second and the language mix.
2. Add `app.bsky.feed.repost` to `WANTED_COLLECTIONS` and measure the
   repost:post ratio.
3. Estimate the deletion rate: collect post creates for five minutes,
   wait an hour, and re-fetch a sample with `bs_get_posts()`.
4. Extend `build_reply()` with a "stop" kill switch that persists a
   do-not-reply list.
5. Compute the Gini coefficient of events per author hash — how
   concentrated is live activity?

## Research project ideas

* **The lifetime of a post.** Stream post creates, then measure
  engagement accumulation and deletion over the following week — a
  prospective cascade study without selection on success.
* **Bots in the open.** Identify disclosed bots (profile keywords),
  stream their activity, and characterize the automated layer of the
  network's activity mix.

## References

### Scholarship

* Morstatter, F., Pfeffer, J., Liu, H., & Carley, K. M. (2013). Is the sample good enough? Comparing data from Twitter's streaming API with Twitter's firehose. *ICWSM*.
* Almuhimedi, H., Wilson, S., Liu, B., Sadeh, N., & Acquisti, A. (2013). Tweets are forever: A large-scale quantitative analysis of deleted tweets. *CSCW '13*.
* Veale, T., & Cook, M. (2018). *Twitterbots: Making Machines that Make Meaning*. MIT Press.
* Gorwa, R., & Guilbeault, D. (2020). Unpacking the social media bot: A typology to guide research and policy. *Policy & Internet*, 12(2).

### Documentation

* Jetstream: <https://github.com/bluesky-social/jetstream>
* Firehose: <https://docs.bsky.app/docs/advanced-guides/firehose>
* Rate limits (points): <https://docs.bsky.app/docs/advanced-guides/rate-limits>
* websocket for R: <https://rstudio.github.io/websocket/>
* bskyr: <https://christophertkenny.com/bskyr/>
