---
title: "Part 08 — LLMs as Research Assistants: Classification and Annotation at Scale"
subtitle: "Bluesky Data Science in R"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

This document is the R counterpart of the Python notebook
[Part 08 - LLMs as Research Assistants](../../Notebooks/Part%2008%20-%20LLMs%20as%20Research%20Assistants/Part%2008%20-%20LLMs%20as%20Research%20Assistants.ipynb).
Content analysis at scale used to require teams of trained coders; large
language models now offer to do the labeling for pennies. Whether their
labels are *defensible as research measures* is an empirical question
that this part answers the traditional way: a codebook, a gold-standard
sample, agreement metrics, and an honest baseline.

### Learning objectives

* Construct a codebook with categories, definitions, and edge rules.
* Build an LLM annotation prompt with structured (JSON) output.
* Always compare against a transparent baseline (keyword rules).
* Compute accuracy, Cohen's kappa, confusion matrices, and per-class
  precision/recall.
* Read disagreements, because that is where the methodology lives.

## Setup

```{r libraries}
library(bskyr)
library(tidyverse)
library(jsonlite)
library(lubridate)
```

As in every part of this series, the shared helper `bsky_client.R` looks
for credentials (an `atproto.json` file or bskyr's environment
variables), authenticates if it finds them, and sets `bsky_authed`. The
`knitr::opts_chunk$set(eval = bsky_authed)` line then makes that flag the
default for the whole document: with credentials every chunk runs;
without them the document still knits, but the API chunks are skipped.

```{r connect}
source("bsky_client.R")
knitr::opts_chunk$set(eval = bsky_authed)
```

## 1. Collecting a corpus of posts

One account's original, non-reply posts — short, varied, and exactly the
kind of material a topic codebook is built for:

```{r corpus}
ACTOR <- "nytimes.com"

feed <- bs_get_author_feed(ACTOR, limit = 200)
if ("reason_type" %in% names(feed)) {
  feed <- filter(feed, is.na(reason_type))
}
if ("reply_parent_uri" %in% names(feed)) {
  feed <- filter(feed, is.na(reply_parent_uri))
}

posts <- feed |>
  transmute(
    uri,
    created_at = map_chr(record, "created_at", .default = NA),
    text = map_chr(record, "text", .default = ""),
    like_count, repost_count, reply_count
  ) |>
  filter(str_squish(text) != "") |>
  slice_head(n = 150)

dir.create("data", showWarnings = FALSE)
write_csv(posts, "data/posts.csv")
write_json(
  list(actor = ACTOR, endpoint = "app.bsky.feed.getAuthorFeed",
       reposts_excluded = TRUE, replies_excluded = TRUE,
       n_collected = nrow(posts),
       collected_at = format(now("UTC"), "%Y-%m-%dT%H:%M:%SZ")),
  "data/collection_metadata.json", pretty = TRUE, auto_unbox = TRUE)

nrow(posts)
```

## 2. Constructing a codebook

A codebook is the *measurement instrument*: mutually exclusive,
collectively exhaustive categories with decision rules a stranger could
apply. Ours mirrors the Python notebook's three-way scheme:

```{r codebook}
LABELS <- c("politics", "science_tech", "other")

CODEBOOK <- '
You are annotating public social media posts with exactly one topic label.

Categories:
- politics: elections, parties, politicians, public policy, government,
  political events and commentary, civic institutions.
- science_tech: research, universities, data science, technology
  platforms and tools, software, scientific findings, academia.
- other: personal life, sports, food, humor with no political or
  scientific referent, and anything that fits neither category above.

Rules:
- Choose ONE label. If a post mixes categories, choose the dominant one.
- Judge only the text given; do not guess at links you cannot see.
- Respond with JSON: {"label": "<one of politics|science_tech|other>",
  "confidence": <0 to 1>}
'
cat(CODEBOOK)
```

## 3. Gold-standard labels

Validation requires labels we trust: a human reads a random sample and
applies the codebook. The Python series hand-coded 60 posts from this
account and committed them with the repository, and because they live in
a plain CSV, the R series can validate against the *same* gold standard
— a nice illustration of why archives beat live collections for
reproducibility:

```{r gold}
gold <- read_csv(
  file.path("..", "..", "Notebooks",
            "Part 08 - LLMs as Research Assistants", "data",
            "gold_labels.csv"),
  show_col_types = FALSE
) |>
  mutate(label_rationale = replace_na(label_rationale, ""))

count(gold, gold_label)
```

```{r gold-rationales}
gold |>
  filter(label_rationale != "") |>
  transmute(text = str_trunc(text, 90), gold_label, label_rationale) |>
  head(4)
```

How many of the gold posts are still in today's live corpus? Posts get
deleted; gold standards should record the text they coded:

```{r gold-overlap}
sum(gold$uri %in% posts$uri)
```

## 4. LLM annotation with the Claude API

### 4.1 The prompt

A per-post prompt pairs the codebook with one post and demands JSON:

```{r prompt}
build_prompt <- function(codebook, post_text) {
  paste0(codebook, "\nPost to annotate:\n\"\"\"\n", post_text, "\n\"\"\"")
}

cat(str_trunc(build_prompt(CODEBOOK, gold$text[[1]]), 700))
```

### 4.2 The annotation loop

The Claude API is an ordinary HTTP API, so
[httr2](https://httr2.r-lib.org/) talks to it directly: build a request
with the authentication headers, attach the JSON body (model, token
budget, and our prompt as a user message), perform it, and parse the
JSON response. Two robustness pieces surround the call:
`parse_label_response()` digs the JSON out of the model's reply — models
sometimes wrap it in prose or code fences, and sometimes invent labels
outside the codebook, both of which must map to `NA`, not crash — and
the half-second sleep keeps us inside API rate limits. Models are listed
at <https://docs.claude.com/en/docs/about-claude/models>:

```{r annotator}
parse_label_response <- function(raw) {
  json <- str_extract(raw, "\\{[^}]*\\}")
  parsed <- tryCatch(fromJSON(json), error = function(e) NULL)
  label <- parsed$label
  if (is.null(label) || !label %in% LABELS) {
    return(list(label = NA_character_, confidence = NA_real_))
  }
  list(label = label, confidence = parsed$confidence %||% NA_real_)
}

annotate_posts_llm <- function(texts, model = "claude-haiku-4-5") {
  map(texts, \(text) {
    resp <- httr2::request("https://api.anthropic.com/v1/messages") |>
      httr2::req_headers(
        "x-api-key" = Sys.getenv("ANTHROPIC_API_KEY"),
        "anthropic-version" = "2023-06-01"
      ) |>
      httr2::req_body_json(list(
        model = model,
        max_tokens = 100,
        messages = list(list(role = "user",
                             content = build_prompt(CODEBOOK, text)))
      )) |>
      httr2::req_perform() |>
      httr2::resp_body_json()
    Sys.sleep(0.5)
    parse_label_response(resp$content[[1]]$text)
  }) |> bind_rows()
}
```

### 4.3 What will this cost?

Annotation cost is input tokens × price. At roughly 4 characters per
token, 60 posts under this codebook are ~25k input tokens — fractions
of a dollar on a small model. Estimate before you run; budget before
you scale.

```{r cost}
est_tokens <- sum(nchar(map_chr(gold$text, \(t) build_prompt(CODEBOOK, t)))) / 4
round(est_tokens)
```

### 4.4 Running the annotator (guarded)

Calls leave your machine and cost money, so the chunk runs only when an
API key is present:

```{r run-llm}
if (nzchar(Sys.getenv("ANTHROPIC_API_KEY"))) {
  llm_labels <- annotate_posts_llm(gold$text)
  gold$llm_label <- llm_labels$label
  write_csv(bind_cols(select(gold, uri, text), llm_labels),
            "data/llm_labels.csv")
} else {
  message("No ANTHROPIC_API_KEY set, so no posts were sent to an API. ",
          "The validation below uses the keyword baseline instead.")
}
```

## 5. A baseline annotator that always runs

Never evaluate an exciting method without a boring one. Keyword rules
are transparent, free, and embarrassingly often competitive. The
function counts how many keywords from each category's list appear in
the lowercased text and picks the category with the most hits, falling
back to `other` when nothing matches — every decision it makes is
auditable by reading 20 lines:

```{r baseline}
KEYWORD_RULES <- list(
  politics = c("election", "vote", "senate", "congress", "policy",
               "democrat", "republican", "trump", "biden", "court",
               "government", "political", "campaign", "ballot"),
  science_tech = c("data", "research", "science", "universit", "paper",
                   "study", "ai", "model", "code", "software", "api",
                   "network", "compute", "course", "student", "bluesky")
)

keyword_label <- function(text) {
  text <- str_to_lower(text)
  hits <- map_int(KEYWORD_RULES, \(words) sum(str_detect(text, fixed(words))))
  if (max(hits) == 0) "other" else names(which.max(hits))
}

gold <- gold |>
  mutate(baseline_label = map_chr(text, keyword_label))

count(gold, baseline_label)
write_csv(select(gold, uri, text, gold_label, baseline_label),
          "data/baseline_predictions.csv")
```

## 6. Validation: how good is the automated annotator?

### 6.1 Accuracy, percent agreement, and Cohen's kappa

Accuracy alone flatters annotators on imbalanced data. Cohen's kappa
corrects for chance agreement — the difference between a coder and a
coin that knows the base rates:

```{r metrics}
cohen_kappa <- function(truth, pred, labels) {
  truth <- factor(truth, levels = labels)
  pred <- factor(pred, levels = labels)
  cm <- table(truth, pred)
  n <- sum(cm)
  po <- sum(diag(cm)) / n
  pe <- sum(rowSums(cm) * colSums(cm)) / n^2
  (po - pe) / (1 - pe)
}

annotator <- if ("llm_label" %in% names(gold)) "llm_label" else "baseline_label"
pred <- gold[[annotator]]

tibble(
  annotator = annotator,
  accuracy = mean(gold$gold_label == pred),
  kappa = cohen_kappa(gold$gold_label, pred, LABELS)
)
```

Benchmarks: kappa above ~0.8 is conventionally "almost perfect," 0.6–0.8
"substantial" — but report the number, not the adjective.

### 6.2 Confusion matrix

Where do the errors live? A confusion matrix cross-tabulates gold labels
(rows) against machine labels (columns); the diagonal is agreement, and
each off-diagonal cell is a specific *kind* of mistake:

```{r confusion}
table(gold = factor(gold$gold_label, LABELS),
      machine = factor(pred, LABELS))
```

### 6.3 Per-category precision and recall

Overall numbers hide category-level failures, so compute, per category:
**precision** (when the machine says "politics," how often is it right?)
and **recall** (of the truly political posts, how many did it find?),
plus their harmonic mean **F1**:

```{r per-class}
per_class <- map(LABELS, \(label) {
  tp <- sum(gold$gold_label == label & pred == label)
  fp <- sum(gold$gold_label != label & pred == label)
  fn <- sum(gold$gold_label == label & pred != label)
  tibble(category = label,
         precision = tp / max(tp + fp, 1),
         recall = tp / max(tp + fn, 1)) |>
    mutate(f1 = 2 * precision * recall / max(precision + recall, 1e-9))
}) |> list_rbind()

per_class

per_class |>
  mutate(validated_at = format(today()), annotator = annotator) |>
  write_csv("data/agreement_metrics.csv")
```

## 7. Error analysis: read the disagreements

Aggregate metrics tell you *how much* the annotator errs; only reading
the disagreements tells you *how* — systematically or randomly, on edge
cases or on the codebook's blind spots:

```{r disagreements}
gold |>
  filter(gold_label != .data[[annotator]]) |>
  transmute(text = str_trunc(text, 110),
            gold_label, machine = .data[[annotator]]) |>
  head(5)
```

## 8. When are LLM labels defensible as research measures?

The emerging consensus (Gilardi et al. 2023; Ziems et al. 2024; Ollion
et al. 2024): LLM annotation is defensible when (1) validated against
human gold labels *on your own task and corpus*, (2) accompanied by the
exact prompt and model version, (3) re-validated when either changes,
and (4) reported with the same agreement statistics you would demand of
human coders. It is not defensible as an oracle. The workflow in this
document *is* the argument: instrument, gold standard, baseline,
metrics, error analysis.

## Ethics and limitations

Sending posts to a commercial API is a *disclosure* of user content to a
third party — minimize (send text only, never handles or DIDs), check
the provider's retention policy, and consider whether your IRB considers
this sharing. LLMs encode training biases that can label minority
dialects or stances systematically differently; per-class metrics and
error reading are your detection tools. Our gold labels were coded by a
single human coder (the author); a real study needs double coding with
inter-coder reliability reported before the gold standard itself can be
trusted.

## Exercises

1. Draw your own random sample of 30 posts, hand-label it with the
   codebook, and compute *your* agreement with the committed gold labels
   on any overlapping posts — inter-coder reliability in miniature.
2. Add a fourth category to the codebook and rerun. Which metrics
   degrade, and why does adding categories usually lower kappa?
3. Write a second keyword baseline with different word lists and measure
   baseline-to-baseline agreement — a floor for what "agreement" means.
4. If you have an API key: run the LLM twice at temperature defaults and
   measure its self-agreement (test-retest reliability).
5. Compute kappa weighted by the LLM's reported confidence. Is the model
   better-calibrated on posts it claims to be sure about?

## Research project ideas

* **Annotator drift audit.** Re-run the same prompt and gold sample
  monthly across model versions and chart agreement over time — a
  reliability study of the instrument itself.
* **Dialect fairness.** Build gold samples in two language varieties and
  test whether LLM agreement differs — extending the algorithmic-bias
  audit tradition to annotation.

## References

### Scholarship

* Gilardi, F., Alizadeh, M., & Kubli, M. (2023). ChatGPT outperforms crowd workers for text-annotation tasks. *PNAS*, 120(30).
* Ziems, C., Held, W., Shaikh, O., Chen, J., Zhang, Z., & Yang, D. (2024). Can large language models transform computational social science? *Computational Linguistics*, 50(1), 237–291.
* Ollion, É., Shen, R., Macanovic, A., & Chokri, A. (2024). The dangers of using proprietary LLMs for research. *Nature Machine Intelligence*, 6, 4–5.
* Krippendorff, K. (2018). *Content Analysis: An Introduction to Its Methodology* (4th ed.). SAGE.
* Cohen, J. (1960). A coefficient of agreement for nominal scales. *Educational and Psychological Measurement*, 20(1), 37–46.

### Documentation

* Claude API: <https://docs.claude.com/en/api/messages>
* httr2: <https://httr2.r-lib.org/>
* bskyr: <https://christophertkenny.com/bskyr/>
