---
title: "Part 06 — Natural Language Processing of Post Content"
subtitle: "Bluesky Data Science in R"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

This document is the R counterpart of the Python notebook
[Part 06 - Natural Language Processing](../../Notebooks/Part%2006%20-%20Natural%20Language%20Processing/Part%2006%20-%20Natural%20Language%20Processing.ipynb).
The methods map onto R's tidy text ecosystem: where Python used
scikit-learn vectorizers and VADER, we use
[tidytext](https://www.tidytextmining.com/) (Silge & Robinson) for
tokenization and sentiment lexicons and
[topicmodels](https://cran.r-project.org/package=topicmodels) for LDA.
The analytical questions are identical: what does a corpus of posts talk
about, in what tone, and how would we know if our measurements were wrong?

### Learning objectives

* Build a multi-account corpus from author feeds.
* Tokenize short social text and weigh what stopword removal throws away.
* Compute word/bigram frequencies and per-account distinctive terms.
* Score sentiment with a lexicon and know where lexicons break.
* Fit and read an LDA topic model, and validate everything.

## Setup

```{r libraries}
library(bskyr)
library(tidyverse)
library(jsonlite)
library(lubridate)
library(tidytext)
library(topicmodels)
```

As in every part of this series, the shared helper `bsky_client.R` looks
for credentials (an `atproto.json` file or bskyr's environment
variables), authenticates if it finds them, and sets `bsky_authed`. The
`knitr::opts_chunk$set(eval = bsky_authed)` line then makes that flag the
default for the whole document: with credentials every chunk runs;
without them the document still knits, but the API chunks are skipped.

```{r connect}
source("bsky_client.R")
knitr::opts_chunk$set(eval = bsky_authed)
```

## 1. Building a corpus from author feeds

### 1.1 A seed account and one snowball wave

We reuse Part 02's design in miniature: collect the seed's posts, find
the accounts it mentions most, and add their posts to form a small
*panel corpus*. The fetcher below should look familiar by now — drop
reposts, pull `text`, `created_at`, `langs`, and `facets` out of the
`record` list-column — with `transmute()` (a `mutate()` that keeps only
the columns it names) producing exactly the tidy frame the rest of the
pipeline needs:

```{r corpus-seed}
SEED_ACCOUNT <- "nytimes.com"
COLLECTED_ON <- format(today(), "%Y-%m-%d")

fetch_posts <- function(actor, max_posts) {
  feed <- bs_get_author_feed(actor, limit = max_posts)
  if ("reason_type" %in% names(feed)) {
    feed <- filter(feed, is.na(reason_type))
  }
  feed |>
    transmute(
      uri,
      author = author_handle,
      created_at = map_chr(record, "created_at", .default = NA),
      text = map_chr(record, "text", .default = ""),
      langs = map(record, \(r) unlist(r$langs)),
      facets = map(record, \(r) r$facets[[1]])
    )
}

seed_posts <- fetch_posts(SEED_ACCOUNT, 600)
nrow(seed_posts)
```

Next, the one-wave snowball. The nested `map()` calls walk the facet
structure exactly as in Part 02 (facets → features → mention DIDs);
`table()` tallies how often each account is mentioned; and we keep the
six most-mentioned accounts that actually have posts:

```{r corpus-panel}
mention_dids <- seed_posts$facets |>
  map(\(facets) {
    map(facets, \(facet) {
      map_chr(facet$features, \(f) {
        if (endsWith(f$`$type`, "#mention")) f$did else NA_character_
      })
    })
  }) |>
  unlist() |>
  discard(is.na)

top_mentioned <- names(sort(table(mention_dids), decreasing = TRUE))[1:8]
profiles <- bs_get_profile(top_mentioned)
panel_handles <- profiles |>
  filter(posts_count > 0) |>
  slice_head(n = 6) |>
  pull(handle)

panel_posts <- map(panel_handles, \(h) {
  out <- fetch_posts(h, 150)
  Sys.sleep(0.1)
  out
}) |> list_rbind()

c(seed = nrow(seed_posts), panel = nrow(panel_posts))
```

### 1.2 Tidying and filtering by language

Text methods below assume English, so we keep posts whose record
*declares* English in `langs` (a self-report by the posting client — a
measurement choice worth remembering), drop empty texts, and
de-duplicate by URI:

```{r corpus-tidy}
corpus <- bind_rows(seed_posts, panel_posts) |>
  distinct(uri, .keep_all = TRUE) |>
  mutate(
    created_at = ymd_hms(created_at, quiet = TRUE),
    english = map_lgl(langs, \(l) "en" %in% l)
  ) |>
  filter(english, str_squish(text) != "") |>
  select(uri, author, created_at, text)

corpus |> count(author, sort = TRUE)
```

## 2. Preprocessing short social text

`unnest_tokens()` lowercases and splits on word boundaries; we then
strip URLs and platform syntax and remove stopwords. Custom stopwords
matter on social text — handles, "rt"-like artifacts, and bare domains
are frequent but meaningless:

```{r tokens}
custom_stop <- tibble(word = c("https", "t.co", "bsky.app", "amp",
                               "gt", "lt", "1", "2", "3"))

tokens <- corpus |>
  mutate(text_clean = text |>
           str_remove_all("https?://\\S+") |>
           str_remove_all("@[\\w.]+")) |>
  unnest_tokens(word, text_clean) |>
  anti_join(stop_words, by = "word") |>
  anti_join(custom_stop, by = "word") |>
  filter(str_detect(word, "[a-z]"))

c(posts = n_distinct(tokens$uri), tokens = nrow(tokens),
  vocabulary = n_distinct(tokens$word))
```

### What did we just throw away?

Stopword lists built for documents amputate meaning from 300-character
posts: negations ("not"), pronouns, and intensifiers all carry signal
for some questions. Print a before/after pair whenever you change the
pipeline, and keep the raw `text` column forever.

```{r before-after}
example <- corpus |> filter(author == SEED_ACCOUNT) |>
  slice_head(n = 1)
example$text
tokens |> filter(uri == example$uri) |> pull(word)
```

## 3. Word and n-gram frequencies

The simplest content analysis is a word count. Single words (*unigrams*)
show vocabulary; adjacent pairs (*bigrams*, built by `unnest_tokens()`
with `token = "ngrams", n = 2`) recover phrases that unigrams shred,
like "data science". `reorder_within()` is a tidytext convenience that
sorts the bars within each facet:

```{r frequencies, fig.width=12, fig.height=6}
unigrams <- tokens |> count(word, sort = TRUE)

bigrams <- corpus |>
  mutate(text_clean = str_remove_all(text, "https?://\\S+")) |>
  unnest_tokens(bigram, text_clean, token = "ngrams", n = 2) |>
  separate(bigram, c("w1", "w2"), sep = " ") |>
  filter(!w1 %in% stop_words$word, !w2 %in% stop_words$word) |>
  unite(bigram, w1, w2, sep = " ") |>
  count(bigram, sort = TRUE)

bind_rows(
  unigrams |> slice_head(n = 20) |> transmute(term = word, n, kind = "unigrams"),
  bigrams |> slice_head(n = 20) |> transmute(term = bigram, n, kind = "bigrams")
) |>
  ggplot(aes(n, reorder_within(term, n, kind))) +
  geom_col(aes(fill = kind), show.legend = FALSE) +
  scale_y_reordered() +
  facet_wrap(~kind, scales = "free") +
  labs(x = "Occurrences in corpus", y = NULL,
       title = paste("Most frequent terms across", nrow(corpus), "posts")) +
  theme_minimal()
```

### 3.1 What makes each account distinctive?

Raw frequency rewards verbosity: the seed account posts the most, so it
dominates every top-words list. *Log-odds ratios* fix this by comparing
each account's rate of using a word against everyone else's rate of
using it — a word scores high for an account when that account uses it
disproportionately, not merely often. The +0.5 *smoothing* keeps words
one group never uses from producing log(0):

```{r log-odds}
account_words <- tokens |> count(author, word)

distinctive <- account_words |>
  group_by(word) |>
  mutate(total = sum(n)) |>
  group_by(author) |>
  mutate(
    own_total = sum(n),
    rest = total - n,
    rest_total = sum(account_words$n) - own_total,
    log_odds = log((n + 0.5) / (own_total - n + 0.5)) -
               log((rest + 0.5) / (rest_total - rest + 0.5))
  ) |>
  filter(n >= 3) |>
  slice_max(log_odds, n = 8) |>
  summarise(distinctive_words = paste(word, collapse = ", "),
            .groups = "drop")

distinctive
```

## 4. Sentiment analysis with a lexicon

Python used VADER, a lexicon tuned for social media. tidytext ships the
Bing lexicon (positive/negative word lists); the mechanics — and the
failure modes — are the same. First, the demos that show what
lexicon-based scoring can and cannot see:

```{r sentiment-demo}
score_text <- function(texts) {
  scores <- tibble(doc = seq_along(texts), text = texts) |>
    unnest_tokens(word, text) |>
    inner_join(get_sentiments("bing"), by = "word") |>
    mutate(value = if_else(sentiment == "positive", 1L, -1L)) |>
    group_by(doc) |>
    summarise(score = sum(value))
  tibble(doc = seq_along(texts)) |>
    left_join(scores, by = "doc") |>
    mutate(score = coalesce(score, 0L)) |>
    pull(score)
}

demos <- c("The new feature is good.",
           "The new feature is not good.",        # negation: NOT handled by Bing
           "Oh great, ANOTHER outage. Just what I needed today.")  # sarcasm: no

tibble(text = demos, score = score_text(demos))
```

Note the failures: unlike VADER, a plain word lexicon misses negation
entirely, and no lexicon catches sarcasm. Validation (Section 7) is not
optional.

```{r sentiment-corpus, fig.width=8, fig.height=5}
corpus <- corpus |> mutate(sentiment = score_text(text))

ggplot(corpus, aes(sentiment)) +
  geom_histogram(binwidth = 1, fill = "mediumpurple", color = "white") +
  geom_vline(xintercept = 0, linewidth = 0.3) +
  labs(x = "Bing sentiment score (positive − negative words)",
       y = "Number of posts",
       title = "Sentiment distribution across the corpus") +
  theme_minimal()
```

### 4.1 Sentiment over time

```{r sentiment-time, fig.width=10, fig.height=5}
corpus |>
  mutate(month = floor_date(created_at, "month")) |>
  group_by(month) |>
  filter(n() >= 20) |>
  summarise(mean_sent = mean(sentiment),
            se = sd(sentiment) / sqrt(n())) |>
  ggplot(aes(month, mean_sent)) +
  geom_ribbon(aes(ymin = mean_sent - 1.96 * se,
                  ymax = mean_sent + 1.96 * se),
              fill = "seagreen", alpha = 0.25) +
  geom_line(color = "seagreen") + geom_point(color = "seagreen") +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.3) +
  labs(x = "Month", y = "Mean sentiment score",
       title = "Monthly mean sentiment across the panel corpus") +
  theme_minimal()
```

### 4.2 Where dictionaries break

Read the extremes — always:

```{r sentiment-extremes}
corpus |>
  filter(author == SEED_ACCOUNT) |>
  arrange(sentiment) |>
  slice(c(1:3, (n() - 2):n())) |>
  transmute(sentiment, text = str_trunc(text, 120))
```

## 5. Topic modeling with LDA

Latent Dirichlet Allocation (LDA) treats each post as a *mixture* of
topics and each topic as a *distribution over words* — and learns both
from word co-occurrence alone, with no labels. The model wants a
**document-term matrix** (rows = posts, columns = words, cells =
counts), which `cast_dtm()` builds from our tidy token counts. We first
drop words appearing in fewer than five posts — they carry almost no
co-occurrence information and balloon the vocabulary — and fix the random
seed, because LDA's fitting is stochastic:

```{r lda}
dtm <- tokens |>
  count(uri, word) |>
  group_by(word) |>
  filter(n_distinct(uri) >= 5) |>          # word in >= 5 posts
  ungroup() |>
  cast_dtm(uri, word, n)

N_TOPICS <- 8
lda_fit <- LDA(dtm, k = N_TOPICS, control = list(seed = 42))

dtm
```

### 5.1 Reading the topics

A fitted LDA model has two faces, and tidytext exposes each as a tidy
tibble. The **beta** matrix gives each word's probability under each
topic — reading the top-12 words per topic is how humans decide what a
topic "is":

```{r topic-words}
topic_words <- tidy(lda_fit, matrix = "beta") |>
  group_by(topic) |>
  slice_max(beta, n = 12) |>
  summarise(top_words = paste(term, collapse = ", "))

topic_words
```

### 5.2 Documents as topic mixtures

The **gamma** matrix is the other face: each post's estimated mixture
over the topics. Here is the single post in the corpus most dominated by
one topic:

```{r gamma}
doc_topics <- tidy(lda_fit, matrix = "gamma")

most_concentrated <- doc_topics |>
  slice_max(gamma, n = 1, with_ties = FALSE)

corpus |>
  filter(uri == most_concentrated$document) |>
  transmute(text = str_trunc(text, 200),
            dominant_topic = most_concentrated$topic,
            share = round(most_concentrated$gamma, 2))
```

### 5.3 Topic prevalence by account

```{r prevalence, fig.width=10, fig.height=5}
doc_topics |>
  left_join(select(corpus, uri, author), by = c(document = "uri")) |>
  group_by(author, topic) |>
  summarise(mean_share = mean(gamma), .groups = "drop") |>
  ggplot(aes(factor(topic), author, fill = mean_share)) +
  geom_tile() +
  geom_text(aes(label = round(mean_share, 2)), size = 3) +
  scale_fill_gradient(low = "white", high = "darkblue") +
  labs(x = "Topic", y = NULL, fill = "Mean share",
       title = "Topic prevalence by account") +
  theme_minimal()
```

### 5.4 Choosing k, and other confessions

We set `k = 8` because it makes a readable demo. There is no true k:
fit several, compare interpretability (and metrics like perplexity or
the [ldatuning](https://cran.r-project.org/package=ldatuning) scores),
and report the choice as a researcher degree of freedom.

## 6. Beyond word counts: transformer sentence embeddings

Modern NLP represents whole sentences as dense vectors that capture
meaning beyond shared vocabulary. From R the practical routes are the
[text](https://r-text.org/) package (which wraps Hugging Face
transformers via reticulate) or calling a Python encoder directly. The
workflow is the same as the Python notebook's: embed posts, then
cluster, classify, or measure similarity in the embedding space.

```{r embeddings, eval=FALSE}
# install.packages("text"); text::textrpp_install()
embeddings <- text::textEmbed(corpus$text[1:3])
dim(embeddings$texts$texts)
```

## 7. Validate, validate, validate

Every measurement above — tokens, sentiment scores, topic labels — is a
*proxy* read off text by rules that do not understand it. The discipline
(Grimmer & Stewart 2013): draw a random sample, read it against the
machine's labels, and quantify the agreement before trusting any
downstream result. Part 08 turns this into a full workflow with gold
labels and agreement metrics.

```{r validation-sample}
set.seed(42)
doc_topics |>
  group_by(document) |>
  slice_max(gamma, n = 1, with_ties = FALSE) |>
  ungroup() |>
  left_join(corpus, by = c(document = "uri")) |>
  filter(author == SEED_ACCOUNT) |>
  slice_sample(n = 8) |>
  transmute(sentiment, dominant_topic = topic,
            text = str_trunc(text, 140))
```

## 8. Saving the corpus and results

```{r save}
dir.create("data", showWarnings = FALSE)

write_csv(corpus, "data/corpus.csv")
write_csv(topic_words, "data/topic_top_words.csv")
write_json(
  list(collected_on = COLLECTED_ON, seed_account = SEED_ACCOUNT,
       panel_accounts = panel_handles, language_filter = "en",
       n_posts_saved = nrow(corpus), lda_topics = N_TOPICS, lda_seed = 42),
  "data/collection_metadata.json", pretty = TRUE, auto_unbox = TRUE)

read_csv("data/corpus.csv", show_col_types = FALSE) |> nrow()
```

## Ethics and limitations

Text is people talking. Quoting strangers' posts verbatim in
publications makes them findable and harassable; paraphrase or aggregate
(our extremes table above is for *your* validation reading, not for
publication). Sentiment lexicons encode the norms of their builders and
miss negation, sarcasm, and community-specific language; LDA topics are
statistical regularities, not meanings, and both will happily produce
confident numbers from garbage. The corpus design (a seed plus its
mention network, English-declared posts only) shapes every count above.

## Exercises

1. Re-run the pipeline keeping stopwords. Which analyses change most?
2. Add the AFINN lexicon (`get_sentiments("afinn")`, via the textdata
   package) and compare scores with Bing on the same posts.
3. Fit LDA with k = 4 and k = 12 and compare the topic-word tables.
   Which is more interpretable?
4. Compute distinctive *bigrams* per account instead of unigrams.
5. Hand-label 30 posts for positive/negative/neutral and compute the
   lexicon's accuracy against your labels.

## Research project ideas

* **Linguistic accommodation.** Using log-odds over time, test whether
  accounts that mention each other converge in vocabulary.
* **Topic dynamics around events.** Fit a topic model to a snowball
  archive (Part 02) spanning a news event and trace topic prevalence
  before, during, and after (combining Parts 05 and 06).

## References

### Scholarship

* Grimmer, J., & Stewart, B. M. (2013). Text as data: The promise and pitfalls of automatic content analysis. *Political Analysis*, 21(3), 267–297.
* Silge, J., & Robinson, D. (2017). *Text Mining with R: A Tidy Approach*. O'Reilly. <https://www.tidytextmining.com/>
* Blei, D. M., Ng, A. Y., & Jordan, M. I. (2003). Latent Dirichlet allocation. *JMLR*, 3, 993–1022.
* Hutto, C., & Gilbert, E. (2014). VADER: A parsimonious rule-based model for sentiment analysis of social media text. *ICWSM*.
* Monroe, B. L., Colaresi, M. P., & Quinn, K. M. (2008). Fightin' words: Lexical feature selection and evaluation for identifying the content of political conflict. *Political Analysis*, 16(4), 372–403.

### Documentation

* tidytext: <https://juliasilge.github.io/tidytext/>
* topicmodels: <https://cran.r-project.org/package=topicmodels>
* text (transformers in R): <https://r-text.org/>
* bskyr: <https://christophertkenny.com/bskyr/>
