---
title: "Part 03 — Social Network Analysis of Follow Relationships"
subtitle: "Bluesky Data Science in R"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

This document is the R counterpart of the Python notebook
[Part 03 - Social Network Analysis](../../Notebooks/Part%2003%20-%20Social%20Network%20Analysis/Part%2003%20-%20Social%20Network%20Analysis.ipynb).
Where the Python version pairs the `atproto` SDK with
[NetworkX](https://networkx.org/), this one pairs
[bskyr](https://christophertkenny.com/bskyr/) with
[igraph](https://r.igraph.org/) — R's workhorse network library — to
crawl and analyze a *1.5-degree ego network* of follow relationships.

### Learning objectives

* Collect follow and follower lists with `bs_get_follows()`/`bs_get_followers()`.
* Confront the boundary specification problem and sample alters honestly.
* Build a directed igraph from crawled edge lists with node attributes.
* Compute degree distributions, centrality, and Louvain communities.
* Persist a network as edge and node tables and reload it faithfully.

## Setup

```{r libraries}
library(bskyr)
library(tidyverse)
library(jsonlite)
library(igraph)
```

As in every part of this series, the shared helper `bsky_client.R` looks
for credentials (an `atproto.json` file or bskyr's environment
variables), authenticates if it finds them, and sets `bsky_authed`. The
`knitr::opts_chunk$set(eval = bsky_authed)` line then makes that flag the
default for the whole document: with credentials every chunk runs;
without them the document still knits, but the API chunks are skipped.

```{r connect}
source("bsky_client.R")
knitr::opts_chunk$set(eval = bsky_authed)
```

## 1. The follow-graph endpoints

A follow is a directed tie: `A → B` means A follows B, and B need not
follow back. Two endpoints expose each side of the relation:

```{r endpoints}
follows_page <- bs_get_follows("nytimes.com", limit = 10)
follows_page |> select(did, handle) |> head(5)

followers_page <- bs_get_followers("nytimes.com", limit = 10)
followers_page$handle
```

A third endpoint, `bs_get_relationships()`, checks ties between specific
pairs — useful when you need a handful of edges, not whole lists.

## 2. Designing the crawl: boundary specification and sampling

Whole-network analysis needs a *boundary*: who is in, who is out
(Laumann, Marsden, & Prensky 1989). The classic compromise is the **ego
network**: one focal account (the *ego*), the accounts it follows (the
*alters*), and all follow ties *among* them — "1.5 degrees" because we
collect the alters' ties to each other but not their ties to outsiders.

```{r ego}
EGO_HANDLE <- "nytimes.com"

ego_profile <- bs_get_profile(EGO_HANDLE)
ego_did <- ego_profile$did

ego_profile |> select(handle, follows_count, followers_count)
```

### Why sample?

Each alter costs at least one request to enumerate *their* follows. We
sample a fixed number of alters with a fixed seed, trading completeness
for a documented, reproducible budget:

```{r sample-alters}
N_ALTERS <- 100
SEED <- 42

ego_follows <- bs_get_follows(ego_did, limit = 4000)

set.seed(SEED)
alters <- ego_follows |> slice_sample(n = min(N_ALTERS, nrow(ego_follows)))

alters$handle |> head(10)
```

## 3. Building the 1.5-degree ego network

The crawl itself: for each alter, fetch up to 200 of their follows, then
keep only those pointing back *into* our node set (`intersect()`). Edges
to accounts outside the boundary are discarded — that is what "1.5
degrees" means. The result of each iteration is a small tibble of
`from → to` pairs, and `list_rbind()` stacks them into one edge list.
We also wrap each fetch in `tryCatch()` because some accounts will be
deleted, suspended, or otherwise unfetchable, and one bad alter should
not kill a 100-account crawl:

```{r crawl}
node_set <- c(ego_did, alters$did)

alter_edges <- map(alters$did, \(did) {
  follows <- tryCatch(bs_get_follows(did, limit = 200),
                      error = function(e) NULL)
  Sys.sleep(0.1)
  if (is.null(follows)) return(NULL)
  tibble(from = did, to = intersect(follows$did, node_set))
}) |> list_rbind()

ego_edges <- tibble(from = ego_did, to = alters$did)
edges <- bind_rows(ego_edges, alter_edges) |>
  filter(from != to) |>
  distinct()

G <- graph_from_data_frame(edges,
                           vertices = tibble(name = node_set),
                           directed = TRUE)

c(nodes = vcount(G), edges = ecount(G))
```

### Node attributes with batched `bs_get_profile()`

A graph of bare DIDs is hard to interpret, so hydrate every node with
profile attributes, 25 per request (Part 01's batching trick). The
`match()` call lines the profile rows up with igraph's vertex order —
igraph stores vertices in its own sequence, so we must never assume it
matches our tibble's row order:

```{r hydrate}
profiles <- map(split(node_set, ceiling(seq_along(node_set) / 25)),
                \(batch) {
                  out <- bs_get_profile(batch)
                  Sys.sleep(0.1)
                  out
                }) |> list_rbind()

idx <- match(V(G)$name, profiles$did)
V(G)$handle          <- profiles$handle[idx]
V(G)$followers_count <- profiles$followers_count[idx]
V(G)$follows_count   <- profiles$follows_count[idx]
V(G)$posts_count     <- profiles$posts_count[idx]

vertex_attr(G, index = ego_did) |> as_tibble()
```

```{r crawl-metadata}
dir.create("data", showWarnings = FALSE)
write_json(
  list(ego_handle = EGO_HANDLE, ego_did = ego_did,
       collected_at = format(Sys.time(), tz = "UTC",
                             "%Y-%m-%dT%H:%M:%SZ"),
       ego_follows_at_collection = nrow(ego_follows),
       n_alters_sampled = N_ALTERS, sampling_seed = SEED),
  "data/collection_metadata.json", pretty = TRUE, auto_unbox = TRUE
)
```

## 4. Describing the network

```{r describe}
tibble(
  nodes = vcount(G),
  edges = ecount(G),
  density = edge_density(G),
  reciprocity = reciprocity(G)
)
```

Density is the share of possible ties that exist; reciprocity is the
share of ties that are returned. Follow networks typically show much
higher reciprocity than chance — following back is a norm.

### Degree distributions

In-degree counts followers *within the network*; out-degree counts
follows. Heavy right tails are the signature of attention inequality:

```{r degrees, fig.width=10, fig.height=4}
deg <- tibble(
  did = V(G)$name,
  handle = V(G)$handle,
  in_degree = degree(G, mode = "in"),
  out_degree = degree(G, mode = "out")
)

deg |>
  pivot_longer(ends_with("degree"), names_to = "direction") |>
  ggplot(aes(value)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "white") +
  facet_wrap(~direction) +
  labs(x = "Degree", y = "Accounts",
       title = "Degree distributions in the ego network") +
  theme_minimal()
```

```{r degrees-loglog, fig.width=6, fig.height=5}
deg |>
  pivot_longer(ends_with("degree"), names_to = "direction") |>
  count(direction, value) |>
  filter(value > 0) |>
  ggplot(aes(value, n, color = direction, shape = direction)) +
  geom_point(alpha = 0.7) +
  scale_x_log10() + scale_y_log10() +
  labs(x = "Degree", y = "Number of accounts",
       title = "Degree distributions, log-log axes") +
  theme_minimal()
```

## 5. Centrality: who matters, and in what sense?

Different centralities answer different questions: in-degree measures
local audience, betweenness measures brokerage between otherwise distant
parts of the network, PageRank measures recursively-weighted endorsement.

```{r centrality}
centrality <- deg |>
  mutate(
    betweenness = betweenness(G, directed = TRUE),
    pagerank = page_rank(G)$vector
  )

centrality |>
  filter(did != ego_did) |>
  arrange(desc(pagerank)) |>
  select(handle, in_degree, betweenness, pagerank) |>
  head(10)
```

The three rankings agree at the top and diverge below — high-betweenness
accounts that are not high in-degree are the bridges an audience count
would miss.

## 6. Community detection

Do the alters cluster into groups? *Community detection* looks for
partitions of the nodes with many ties inside groups and few between
them. Louvain modularity maximization is the standard fast algorithm; it
is defined for undirected graphs, so we collapse direction first.
*Modularity* (the quality score it maximizes) runs from 0 (no more
internal ties than chance) toward 1 (perfectly separated clusters):

```{r communities}
G_undirected <- as_undirected(G, mode = "collapse")

set.seed(SEED)
communities <- cluster_louvain(G_undirected)

V(G)$community <- membership(communities)

c(n_communities = length(communities),
  modularity = round(modularity(communities), 3))
```

```{r community-members}
nodes_df <- tibble(
  did = V(G)$name,
  handle = V(G)$handle,
  followers_count = V(G)$followers_count,
  community = V(G)$community
)

nodes_df |>
  add_count(community, name = "community_size") |>
  group_by(community, community_size) |>
  slice_max(followers_count, n = 3, with_ties = FALSE) |>
  summarise(top_members = paste(handle, collapse = ", "), .groups = "drop") |>
  arrange(desc(community_size)) |>
  head(4)
```

Reading a few high-profile members per community usually reveals an
interpretable social context — a discipline, a workplace, a hobby. That
interpretive step is yours, not the algorithm's.

## 7. Visualizing the network

A network drawing is a *map*, not a measurement — but it is often the
fastest way to see the community structure we just quantified. The
Fruchterman–Reingold layout pulls connected nodes together; we color
nodes by their Louvain community, size them by in-degree, and label only
the ego. Layouts are random, so `set.seed()` makes the picture
reproducible:

```{r plot, fig.width=11, fig.height=11}
set.seed(SEED)
layout <- layout_with_fr(G_undirected)

plot(G_undirected,
     layout = layout,
     vertex.color = V(G)$community,
     vertex.size = 2 + sqrt(degree(G, mode = "in")),
     vertex.label = ifelse(V(G)$name == ego_did, "ego", NA),
     vertex.label.font = 2,
     edge.color = adjustcolor("gray", alpha.f = 0.3),
     edge.arrow.size = 0,
     main = paste0("1.5-degree ego network of ", EGO_HANDLE,
                   "\ncolored by Louvain community, sized by in-degree"))
```

## 8. Saving and reloading the network

Edge and node tables are the most portable network format:

```{r save}
write_csv(rename(edges, source_did = from, target_did = to),
          "data/edges.csv")
write_csv(nodes_df, "data/nodes.csv")

c(edges = nrow(edges), nodes = nrow(nodes_df))
```

```{r reload}
edges_loaded <- read_csv("data/edges.csv", show_col_types = FALSE)
nodes_loaded <- read_csv("data/nodes.csv", show_col_types = FALSE)

G_loaded <- graph_from_data_frame(
  edges_loaded,
  vertices = rename(nodes_loaded, name = did),
  directed = TRUE
)

c(same_nodes = setequal(V(G_loaded)$name, V(G)$name),
  same_edge_count = ecount(G_loaded) == ecount(G))
```

Any analysis re-runs from disk — e.g., the top PageRank alters:

```{r reload-analysis}
pr <- page_rank(G_loaded)$vector
pr <- pr[names(pr) != ego_did]

tibble(handle = V(G_loaded)$handle[match(names(pr), V(G_loaded)$name)],
       pagerank = pr) |>
  slice_max(pagerank, n = 5)
```

## 9. Research designs with ego networks

Ego networks support comparative designs: collect many egos and model how
network properties (density, brokerage, community diversity) relate to
outcomes (information exposure, support, influence). Classic findings —
Granovetter's weak ties, Burt's structural holes — are statements about
ego-network structure that these data can test at platform scale.

## Ethics and limitations

A follow list is public, but assembling thousands of ties into a map of
someone's social world creates information that never existed in any one
place — treat crawled networks as sensitive even when every edge is
public. Our boundary choices (one ego, 100 sampled alters, two pages of
follows per alter) are *measurement decisions* that shape every statistic
above; report them. Centrality and community labels are descriptions of
the crawled graph, not of the people. And the graph is a snapshot:
follows churn constantly.

## Exercises

1. Re-crawl with `N_ALTERS = 50` and compare density and modularity to
   the 100-alter network — how sensitive are they to the boundary?
2. Add eigenvector centrality (`eigen_centrality()`) to the comparison
   table. Which accounts move most between rankings?
3. Compute transitivity (`transitivity(G_undirected)`) and compare it to
   the density — what does the gap tell you about clustering?
4. Build the ego network from *followers* instead of follows — sampling
   alters from one page of recent followers, since this ego has over a
   million. How does its structure differ?
5. Color the network plot by `followers_count` quantile instead of
   community. Do the big accounts cluster together?

## Research project ideas

* **Homophily in follow ties.** Annotate alters with public profile
  features (e.g. occupation keywords from bios) and test whether ties are
  more likely within categories than between them.
* **Brokerage and information arrival.** For a panel of egos, test
  whether accounts bridging communities (high betweenness) post links
  earlier than accounts embedded in a single community.

## References

### Scholarship

* Laumann, E. O., Marsden, P. V., & Prensky, D. (1989). The boundary specification problem in network analysis. In *Research Methods in Social Network Analysis*.
* Granovetter, M. (1973). The strength of weak ties. *American Journal of Sociology*, 78(6), 1360–1380.
* Burt, R. S. (2004). Structural holes and good ideas. *American Journal of Sociology*, 110(2), 349–399.
* Kwak, H., Lee, C., Park, H., & Moon, S. (2010). What is Twitter, a social network or a news media? *WWW '10*.
* Blondel, V. D., Guillaume, J.-L., Lambiotte, R., & Lefebvre, E. (2008). Fast unfolding of communities in large networks. *Journal of Statistical Mechanics*, P10008.

### Documentation

* igraph for R: <https://r.igraph.org/>
* `app.bsky.graph.getFollows`: <https://docs.bsky.app/docs/api/app-bsky-graph-get-follows>
* bskyr: <https://christophertkenny.com/bskyr/>
* R Markdown: <https://rmarkdown.rstudio.com/docs/>
