---
title: "Part 02 — Searching and Snowballing: Building a Post Archive"
subtitle: "Bluesky Data Science in R"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

This document is the R counterpart of the Python notebook
[Part 02 - Searching and Snowballing](../../Notebooks/Part%2002%20-%20Searching%20and%20Snowballing/Part%2002%20-%20Searching%20and%20Snowballing.ipynb).
Part 01 archived one account we chose in advance. This part builds a
*topical* archive when we don't know in advance whose posts matter, by
two complementary routes: keyword search, and snowball sampling along the
mention and reply ties embedded in posts.

### Learning objectives

* Query `app.bsky.feed.searchPosts` with date, language, and tag filters.
* Read the anatomy of a post record, especially its *facets*.
* Implement snowball sampling over mention and reply-parent edges.
* Diagnose saturation: when does another wave stop adding new material?
* Assemble, deduplicate, and persist a documented post archive.

## Setup

```{r libraries}
library(bskyr)
library(tidyverse)
library(jsonlite)
library(lubridate)
```

As in every part of this series, the shared helper `bsky_client.R` looks
for credentials (an `atproto.json` file or bskyr's environment
variables), authenticates if it finds them, and sets `bsky_authed`. The
`knitr::opts_chunk$set(eval = bsky_authed)` line then makes that flag the
default for the whole document: with credentials every chunk runs;
without them the document still knits, but the API chunks are skipped.

```{r connect}
source("bsky_client.R")
knitr::opts_chunk$set(eval = bsky_authed)
```

One difference from the Python series is worth flagging immediately: in
Python, search is the *one* endpoint in this part that demands
authentication (the public AppView returns 403 for it). bskyr
authenticates every call, so when this document runs at all, search runs
too.

## 1. Searching for posts

`bs_search_posts()` wraps `app.bsky.feed.searchPosts`. A quoted phrase
matches exactly; `since`/`until`, `lang`, and `tag` narrow the window:

```{r search-basic}
hits <- bs_search_posts('"snowball sampling"', limit = 5)

hits |>
  mutate(text = map_chr(record, "text", .default = NA)) |>
  select(author_handle, text)
```

### Paginating a search

Like every list endpoint, search paginates with cursors, and bskyr
follows them automatically when `limit > 100`. Here are the last seven
days of English-language posts tagged `#openscience`:

```{r search-paged}
until <- now("UTC")
since <- until - days(7)

results <- bs_search_posts(
  "#openscience",
  since = format(since, "%Y-%m-%dT%H:%M:%SZ"),
  until = format(until, "%Y-%m-%dT%H:%M:%SZ"),
  lang  = "en",
  sort  = "latest",
  limit = 300
)

c(posts = nrow(results), accounts = n_distinct(results$author_did))
```

### When you cannot search

Search indexes only what the index operator chooses to index, ranks
results opaquely, and (on some platforms) disappears entirely. The
snowball design in Section 4 builds a topical archive from *public graph
structure alone* — a strategy that survives even when search does not.

## 2. The anatomy of a post record

Every post is a record in its author's repository. The *view* of a post
adds engagement counters and author info; the *record* holds what the
author wrote. In bskyr's cleaned tibbles the record is a list-column:

```{r record-anatomy}
feed <- bs_get_author_feed("nytimes.com", limit = 100)

example <- feed |>
  mutate(n_facets = map_int(record, \(r) length(r$facets[[1]]))) |>
  filter(n_facets > 0) |>
  slice_max(n_facets, n = 1, with_ties = FALSE)

example |> mutate(text = map_chr(record, "text")) |> pull(text)
```

### Facets

Mentions, hashtags, and links are not parsed out of the text by clients —
they are declared by the author's client as **facets**: byte ranges of
the text paired with typed *features*. Inspect the example post's facets:

```{r facets}
facets <- example$record[[1]]$facets[[1]]

str(facets, max.level = 3, list.len = 4)
```

Each feature's `$type` ends in `#mention` (with a `did`), `#tag` (with a
`tag`), or `#link` (with a `uri`). Because mentions carry DIDs — stable
account identifiers — facets turn a pile of text into a *network*.

### Languages, replies, and embeds

The record also declares its languages (`langs`), whether it replies to
another post, and any embedded media — all of which later parts use.

## 3. Facet helpers

Three small functions pull the network out of a feed tibble; they are
this part's equivalent of the Python `snowball.py` module. The logic of
`extract_features()`: a record's `facets` element is a list of facets,
each facet holds a list of features, and each feature has a `$type`
ending in `#mention`, `#tag`, or `#link`. We walk that nesting with two
layers of `map()`, keep the features whose type matches, and pull out
the field we want (`did` for mentions, `tag` for hashtags, `uri` for
links). `discard(..., is.na)` then drops the non-matches:

```{r helpers}
extract_features <- function(record, feature_type, field) {
  facets <- record$facets[[1]]
  if (is.null(facets)) return(character(0))
  out <- map(facets, \(facet) {
    map_chr(facet$features, \(feature) {
      if (endsWith(feature$`$type`, feature_type)) feature[[field]]
      else NA_character_
    })
  })
  discard(unlist(out), is.na)
}

extract_mentions <- function(record) extract_features(record, "#mention", "did")
extract_tags     <- function(record) extract_features(record, "#tag", "tag")
extract_links    <- function(record) extract_features(record, "#link", "uri")

example_record <- example$record[[1]]
list(
  mentions = extract_mentions(example_record),
  hashtags = extract_tags(example_record),
  links    = extract_links(example_record)
)
```

## 4. Snowball sampling: from one seed to a corpus

Snowball sampling starts from a *seed*, observes whom the seed engages,
collects those accounts, and repeats. Set the design parameters first —
every cap is a promise to the API and to your own time:

```{r params}
SEED_ACTOR         <- "nytimes.com"
WAVE0_MAX_POSTS    <- 200  # posts to collect from the seed
WAVE1_ACCOUNTS     <- 8    # most-engaged accounts to crawl in wave 1
POSTS_PER_ACCOUNT  <- 100  # posts per snowballed account
MAX_TOTAL_ACCOUNTS <- 15   # hard cap across all waves

COLLECTED_AT <- format(now("UTC"), "%Y-%m-%dT%H:%M:%SZ")
COLLECTED_AT
```

### Wave 0: the seed account

A small fetcher that returns original posts (reposts dropped) with the
columns the snowball needs:

```{r fetch-author}
fetch_author_posts <- function(actor, max_posts = 100) {
  feed <- bs_get_author_feed(actor, limit = max_posts)
  if (!is.null(feed) && "reason_type" %in% names(feed)) {
    feed <- filter(feed, is.na(reason_type))
  }
  feed |>
    mutate(
      text       = map_chr(record, "text", .default = NA),
      created_at = map_chr(record, "created_at", .default = NA),
      is_reply   = if ("reply_parent_uri" %in% names(feed))
                     !is.na(reply_parent_uri) else FALSE,
      reply_parent_did = if ("reply_parent_author_did" %in% names(feed))
                           reply_parent_author_did else NA_character_,
      mentions   = map(record, extract_mentions),
      hashtags   = map(record, extract_tags)
    ) |>
    select(uri, author_did, author_handle, created_at, text, is_reply,
           reply_parent_did, mentions, hashtags,
           like_count, repost_count, reply_count)
}

wave0 <- fetch_author_posts(SEED_ACTOR, WAVE0_MAX_POSTS)
seed_did <- wave0$author_did[[1]]

c(posts = nrow(wave0))
range(wave0$created_at)
```

### Whom does the seed engage?

Whom did those 200 posts engage? We pool every mentioned DID and every
reply-parent DID, tally them with base R's `table()`, and sort so the
most-engaged accounts come first. The top `WAVE1_ACCOUNTS` of them form
the next wave's crawl list — note that we exclude the seed itself, since
we already have its posts:

```{r engaged}
count_engaged <- function(posts, exclude = character(0)) {
  c(unlist(posts$mentions), discard(posts$reply_parent_did, is.na)) |>
    discard(\(did) did %in% exclude) |>
    table() |>
    sort(decreasing = TRUE)
}

engaged0 <- count_engaged(wave0, exclude = seed_did)
wave1_dids <- names(head(engaged0, WAVE1_ACCOUNTS))

bs_get_profile(wave1_dids) |>
  mutate(posts_engaging = as.integer(engaged0[did])) |>
  select(handle, did, posts_engaging)
```

### Wave 1: crawl the frontier

```{r wave1}
crawl_accounts <- function(dids, posts_each) {
  map(dids, \(did) {
    posts <- tryCatch(fetch_author_posts(did, posts_each),
                      error = function(e) NULL)
    Sys.sleep(0.1)
    posts
  }) |> list_rbind()
}

wave1 <- crawl_accounts(wave1_dids, POSTS_PER_ACCOUNT)
crawled <- c(seed_did, wave1_dids)

c(posts = nrow(wave1), accounts = length(wave1_dids))
```

### Wave 2: one more hop

```{r wave2}
engaged1 <- count_engaged(wave1, exclude = crawled)
budget <- MAX_TOTAL_ACCOUNTS - length(crawled)
wave2_dids <- names(head(engaged1, budget))

wave2 <- crawl_accounts(wave2_dids, POSTS_PER_ACCOUNT)
crawled <- c(crawled, wave2_dids)

c(posts = nrow(wave2), total_accounts_crawled = length(crawled))
```

## 5. Saturation diagnostics: when to stop

Each wave should be judged by its *marginal yield*: how many genuinely
new posts, accounts, and hashtags it added. When the curves flatten, the
snowball has saturated the neighborhood of the seed.

One piece of R syntax to flag in the next chunk: the `<<-` operator
assigns to a variable *outside* the current function, which lets our
per-wave function update the running sets of already-seen posts,
accounts, and hashtags as `imap()` walks over the waves in order:

```{r saturation, fig.width=10, fig.height=3.5}
waves <- list(`0` = wave0, `1` = wave1, `2` = wave2)

seen_posts <- seen_accounts <- seen_tags <- character(0)
wave_stats <- imap(waves, \(posts, wave) {
  new_posts <- setdiff(posts$uri, seen_posts)
  accounts  <- unique(c(posts$author_did, unlist(posts$mentions),
                        discard(posts$reply_parent_did, is.na)))
  new_accounts <- setdiff(accounts, seen_accounts)
  tags <- unique(unlist(posts$hashtags))
  new_tags <- setdiff(tags, seen_tags)
  seen_posts    <<- c(seen_posts, new_posts)
  seen_accounts <<- c(seen_accounts, new_accounts)
  seen_tags     <<- c(seen_tags, new_tags)
  tibble(wave = wave, posts_collected = nrow(posts),
         new_posts = length(new_posts),
         new_accounts = length(new_accounts),
         new_hashtags = length(new_tags))
}) |> list_rbind()

wave_stats

wave_stats |>
  pivot_longer(starts_with("new_"), names_to = "yield") |>
  ggplot(aes(wave, value, fill = yield)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~yield, scales = "free_y") +
  labs(x = "Wave", y = "New per wave",
       title = "Marginal yield of each snowball wave") +
  theme_minimal()
```

## 6. Building the archive

Now assemble the deliverable. `list_rbind()` stacks the three waves into
one tibble (tagging each post with the wave that found it),
`distinct(uri, ...)` keeps only the first sighting of each post — the
same post can be collected twice when snowballed accounts reply to each
other — and `ymd_hms()` parses the timestamps:

```{r archive}
archive <- imap(waves, \(posts, wave) mutate(posts, wave = as.integer(wave))) |>
  list_rbind() |>
  distinct(uri, .keep_all = TRUE) |>
  mutate(created_at = ymd_hms(created_at, quiet = TRUE)) |>
  arrange(created_at)

dim(archive)
```

### Saving the archive — and its provenance

List-columns (mentions, hashtags) don't fit CSV, so the full archive goes
to RDS and a flat projection to CSV; the metadata records every design
parameter:

```{r save-archive}
dir.create("data", showWarnings = FALSE)

saveRDS(archive, "data/archive.rds")
archive |>
  mutate(across(c(mentions, hashtags),
                \(col) map_chr(col, paste, collapse = " "))) |>
  write_csv("data/archive.csv")

metadata <- list(
  collected_at = COLLECTED_AT,
  collection_method = "snowball sample over mention and reply-parent edges",
  seed_actor = SEED_ACTOR,
  seed_did = seed_did,
  parameters = list(
    wave0_max_posts = WAVE0_MAX_POSTS,
    wave1_accounts = WAVE1_ACCOUNTS,
    posts_per_account = POSTS_PER_ACCOUNT,
    max_total_accounts = MAX_TOTAL_ACCOUNTS,
    reposts_included = FALSE,
    replies_included = TRUE
  ),
  accounts_crawled = sort(crawled),
  total_posts = nrow(archive)
)
write_json(metadata, "data/archive_metadata.json",
           pretty = TRUE, auto_unbox = TRUE)
```

### Proving the round trip

```{r roundtrip}
reloaded <- readRDS("data/archive.rds")

stopifnot(nrow(reloaded) == nrow(archive),
          all(reloaded$uri == archive$uri))
c(posts = nrow(reloaded), authors = n_distinct(reloaded$author_did))
```

### Re-hydrating archived posts

Archives go stale: posts get deleted, counters keep moving. `bs_get_posts()`
re-fetches a batch of URIs so you can measure attrition:

```{r rehydrate}
set.seed(42)
sample_uris <- sample(reloaded$uri, 10)

fresh <- bs_get_posts(sample_uris)

tibble(uri = sample_uris) |>
  left_join(select(fresh, uri, like_count_now = like_count), by = "uri") |>
  left_join(select(reloaded, uri, like_count), by = "uri") |>
  mutate(still_exists = !is.na(like_count_now)) |>
  select(like_count, like_count_now, still_exists)
```

## 7. A first look at the archive

```{r top-tags, fig.width=8, fig.height=5}
archive |>
  select(hashtags) |>
  unnest(hashtags) |>
  count(hashtags, sort = TRUE) |>
  slice_head(n = 15) |>
  ggplot(aes(n, fct_reorder(hashtags, n))) +
  geom_col(fill = "seagreen") +
  labs(x = "Posts using hashtag", y = NULL,
       title = paste0("Top hashtags in the archive (n=", nrow(archive), " posts)")) +
  theme_minimal()
```

```{r by-author}
archive |>
  group_by(author_handle) |>
  summarise(posts = n(), replies = sum(is_reply),
            median_likes = median(like_count, na.rm = TRUE)) |>
  arrange(desc(posts))
```

## 8. Research design: what kind of sample is this?

A snowball sample is not a probability sample: every inclusion
probability depends on the seed, and well-connected accounts are
oversampled by construction. That makes it excellent for *mapping a
community around a known starting point* and unsuitable for estimating
population frequencies. Respondent-driven sampling (Heckathorn 1997)
shows how far the design can be pushed when you track recruitment chains
carefully; our wave column preserves exactly that information.

## Ethics and limitations

The snowball pulls in *third parties* — people the seed mentioned who
never interacted with us — so minimize what you keep about them and
aggregate wherever possible. Mentions and replies are public performance,
not consent to study. The archive freezes posts people may later delete
(re-hydration showed attrition is real); honor deletions in long-lived
datasets. And remember the design's blind spots: accounts the seed never
engages are invisible, no matter how central they are to the topic.

## Exercises

1. Re-run the snowball with a different seed and compare the top-hashtag
   tables — how much do the archives overlap?
2. Extend `count_engaged()` to weight reply-parents more heavily than
   mentions. Does wave 1 change?
3. Add a third wave. What does the marginal-yield plot say about it?
4. Use `bs_search_posts()` with a `tag` filter to collect the same topic
   by search, and compare author sets with the snowball archive.
5. Measure attrition: re-hydrate 50 archived posts a week after
   collection and estimate the deletion rate.

## Research project ideas

* **Community boundary mapping.** Seed snowballs from three accounts in
  different communities and measure where the resulting archives begin to
  overlap — an empirical map of community boundaries.
* **Search versus snowball coverage.** For a topic with a clear hashtag,
  quantify what fraction of search-retrieved posts the snowball recovers,
  and what the snowball finds that search misses.

## References

### Scholarship

* Heckathorn, D. D. (1997). Respondent-driven sampling: A new approach to the study of hidden populations. *Social Problems*, 44(2), 174–199.
* González-Bailón, S., Wang, N., Rivero, A., Borge-Holthoefer, J., & Moreno, Y. (2014). Assessing the bias in samples of large online networks. *Social Networks*, 38, 16–27.
* Tufekci, Z. (2014). Big questions for social media big data. *ICWSM*, 8(1), 505–514.
* Bruns, A., & Burgess, J. (2011). The use of Twitter hashtags in the formation of ad hoc publics. *ECPR General Conference*.

### Documentation

* `app.bsky.feed.searchPosts`: <https://docs.bsky.app/docs/api/app-bsky-feed-search-posts>
* Richtext facets: <https://docs.bsky.app/docs/advanced-guides/post-richtext>
* bskyr: <https://christophertkenny.com/bskyr/>
* R Markdown: <https://rmarkdown.rstudio.com/docs/>
