---
title: "Part 09 — Algorithmic Curation and Governance: Custom Feeds, Lists, Labelers, and Starter Packs"
subtitle: "Bluesky Data Science in R"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

This document is the R counterpart of the Python notebook
[Part 09 - Algorithmic Curation and Governance](../../Notebooks/Part%2009%20-%20Algorithmic%20Curation%20and%20Governance/Part%2009%20-%20Algorithmic%20Curation%20and%20Governance.ipynb).
On most platforms "the algorithm" is a sealed box that researchers audit
from outside, if at all. Bluesky decomposes curation into inspectable
parts — **feed generators**, **lists**, **starter packs**, and
**labelers** — all addressable through public endpoints. This part runs
a small *algorithm audit*: sampling posts from several custom feeds and
comparing what each one chooses to show.

### Learning objectives

* Enumerate feed generators and understand them as public records.
* Sample posts from custom feeds and build a comparable archive.
* Quantify feed composition: recency, engagement, concentration, media.
* Inspect lists, starter packs, and moderation labels.
* Frame audits of curation systems without platform cooperation.

## Setup

```{r libraries}
library(bskyr)
library(tidyverse)
library(jsonlite)
library(lubridate)
```

As in every part of this series, the shared helper `bsky_client.R` looks
for credentials (an `atproto.json` file or bskyr's environment
variables), authenticates if it finds them, and sets `bsky_authed`. The
`knitr::opts_chunk$set(eval = bsky_authed)` line then makes that flag the
default for the whole document: with credentials every chunk runs;
without them the document still knits, but the API chunks are skipped.

```{r connect}
source("bsky_client.R")
knitr::opts_chunk$set(eval = bsky_authed)

BSKY_APP_DID <- "did:plc:z72i7hdynmk6r22z27h6tvur"  # bsky.app
BSKY_MOD_DID <- "did:plc:ar7c4by46qjdydhdevvrndac"  # moderation.bsky.app
```

## 1. Feed generators are public records

A custom feed is itself a record in its creator's repository — with a
name, a description, and a like count of subscribers. Everything
`bsky.app` publishes:

```{r actor-feeds}
bsky_feeds <- bs_get_feeds("bsky.app", limit = 50)

bsky_feeds |>
  select(display_name, like_count, description) |>
  arrange(desc(like_count))
```

`bs_get_feed_generator()` resolves any feed URI to its metadata, so
audits can name exactly which version of which algorithm they sampled:

```{r generators}
CANDIDATE_FEEDS <- c(
  "Discover" = paste0("at://", BSKY_APP_DID,
                      "/app.bsky.feed.generator/whats-hot"),
  "What's Hot Classic" = paste0("at://", BSKY_APP_DID,
                                "/app.bsky.feed.generator/hot-classic"),
  "Science" = "at://did:plc:jfhpnnst6flqway4eaeqzj2a/app.bsky.feed.generator/for-science",
  "News" = "at://did:plc:kkf4naxqmweop7dv4l2iqqf5/app.bsky.feed.generator/verified-news"
)

generators <- bs_get_feed_generators(unname(CANDIDATE_FEEDS))

generators |>
  select(display_name, creator_handle, like_count)
```

## 2. Sampling posts from custom feeds

`bs_get_feed()` returns what a feed generator is serving right now. Like
Part 07, we take the raw response (`clean = FALSE`) so that whatever
embed types the feeds surface cannot break us, and flatten just the
fields the audit needs. Note the bookkeeping that makes an audit an
audit: every sample records *when it was retrieved*, because a feed is a
moving target and "the Discover feed" without a timestamp is not a
defined object:

```{r fetch-feeds}
feed_posts <- function(uri, target = 150) {
  pages <- tryCatch(bs_get_feed(uri, limit = target, clean = FALSE),
                    error = function(e) NULL)
  if (is.null(pages)) return(NULL)
  items <- list_flatten(map(pages, "feed"))
  map(items, \(item) {
    post <- item$post
    embed_type <- post$embed$`$type` %||% ""
    tibble(
      uri = post$uri,
      author_did = post$author$did,
      author_handle = post$author$handle,
      created_at = post$record$createdAt %||% NA_character_,
      like_count = post$likeCount %||% 0L,
      repost_count = post$repostCount %||% 0L,
      has_media = str_detect(embed_type, "images|video")
    )
  }) |> list_rbind()
}

samples <- imap(CANDIDATE_FEEDS, \(uri, name) {
  retrieved_at <- now("UTC")
  posts <- feed_posts(uri)
  Sys.sleep(0.5)
  if (is.null(posts)) return(NULL)
  mutate(posts, feed = name, retrieved_at = retrieved_at)
})

map_int(compact(samples), nrow)
```

### A no-algorithm baseline

Audits need a reference point. A *list feed* — the reverse-chronological
posts of a fixed roster — has no ranking algorithm at all:

```{r baseline}
bsky_lists <- bs_get_actor_lists("bsky.app", limit = 50)

baseline_list <- bsky_lists |>
  filter(str_detect(purpose, "curatelist")) |>
  slice_head(n = 1)

baseline_posts <- NULL
if (nrow(baseline_list) == 1) {
  pages <- bs_get_list_feed(baseline_list$uri, limit = 150, clean = FALSE)
  items <- list_flatten(map(pages, "feed"))
  baseline_posts <- map(items, \(item) {
    post <- item$post
    embed_type <- post$embed$`$type` %||% ""
    tibble(uri = post$uri, author_did = post$author$did,
           author_handle = post$author$handle,
           created_at = post$record$createdAt %||% NA_character_,
           like_count = post$likeCount %||% 0L,
           repost_count = post$repostCount %||% 0L,
           has_media = str_detect(embed_type, "images|video"))
  }) |> list_rbind() |>
    mutate(feed = "Baseline (list feed)", retrieved_at = now("UTC"))
}

posts_df <- bind_rows(compact(samples), baseline_posts) |>
  mutate(created_at = ymd_hms(created_at, quiet = TRUE),
         age_hours = as.numeric(difftime(retrieved_at, created_at,
                                         units = "hours")))

count(posts_df, feed)
```

```{r save-archive}
dir.create("data", showWarnings = FALSE)
saveRDS(posts_df, "data/feed_posts.rds")
write_csv(posts_df, "data/feed_posts.csv")
```

## 3. Auditing feed composition

For each feed: how fresh are its posts, how engaged, how concentrated on
few authors, how visual? Concentration gets two measures. The
*top-10-author share* is intuitive: what fraction of the feed comes from
its ten most frequent authors? The **Herfindahl–Hirschman index** (HHI),
borrowed from antitrust economics, is the sum of squared author shares:
it approaches 1 when one author dominates and 1/n when posts are spread
evenly over n authors:

```{r composition}
hhi <- function(authors) {
  shares <- table(authors) / length(authors)
  sum(shares^2)
}

comparison <- posts_df |>
  group_by(feed) |>
  summarise(
    posts = n(),
    unique_authors = n_distinct(author_did),
    top10_author_share = {
      top <- head(sort(table(author_did), decreasing = TRUE), 10)
      sum(top) / n()
    },
    hhi = hhi(author_did),
    median_age_hours = median(age_hours, na.rm = TRUE),
    median_likes = median(like_count, na.rm = TRUE),
    media_share = mean(has_media)
  )

write_csv(comparison, "data/feed_comparison.csv")
comparison
```

```{r composition-plots, fig.width=11, fig.height=8}
posts_df |>
  filter(age_hours > 0) |>
  ggplot(aes(age_hours, color = feed)) +
  stat_ecdf() +
  scale_x_log10() +
  labs(x = "Post age at retrieval (hours)",
       y = "Cumulative share of posts",
       title = "Recency profile of each feed") +
  theme_minimal()

posts_df |>
  ggplot(aes(like_count + 1, color = feed)) +
  stat_ecdf() +
  scale_x_log10() +
  labs(x = "Likes at retrieval (+1)", y = "Cumulative share of posts",
       title = "Engagement profile of each feed") +
  theme_minimal()
```

### What the differences reveal

Ranked discovery feeds serve recent, high-engagement posts from many
authors; topical feeds re-surface older niche material; the list
baseline simply mirrors its roster's posting volume, concentration and
all. None of this required platform cooperation — only public endpoints
and a sampling design.

## 4. Audience overlap between feeds

Do different feeds show the same *accounts*? **Jaccard similarity** —
the size of the intersection of two author sets divided by the size of
their union — runs from 0 (no shared authors) to 1 (identical author
sets). `crossing()` builds every pair of feeds, and `pivot_wider()`
arranges the result as a matrix:

```{r jaccard}
author_sets <- posts_df |>
  group_by(feed) |>
  summarise(authors = list(unique(author_did)))

jaccard <- function(a, b) length(intersect(a, b)) / length(union(a, b))

crossing(f1 = author_sets$feed, f2 = author_sets$feed) |>
  mutate(jaccard = map2_dbl(
    f1, f2,
    \(x, y) jaccard(author_sets$authors[[match(x, author_sets$feed)]],
                    author_sets$authors[[match(y, author_sets$feed)]])
  )) |>
  pivot_wider(names_from = f2, values_from = jaccard)
```

Near-zero overlaps mean the feeds are sampling almost disjoint slices of
the network — each algorithm is a different *editorial regime*.

## 5. Lists: shared rosters for curation and moderation

Lists are public records too — rosters that anyone can subscribe to as a
feed, or (for modlists) as a mute/block set:

```{r lists}
bsky_lists |>
  transmute(name, purpose = str_extract(purpose, "[^#]+$"),
            members = list_item_count,
            description = str_trunc(replace_na(description, ""), 80))

if (nrow(baseline_list) == 1) {
  members <- bs_get_list(baseline_list$uri, limit = 200)
  members |> select(subject_handle, subject_display_name) |> head(10)
}
```

## 6. Starter packs: curating the network itself

Starter packs bundle accounts (and feeds) so newcomers can follow an
entire community at once — curation of the *graph*, not of content:

```{r starter-packs}
packs <- bs_get_actor_starter_packs("bsky.app", limit = 50)

packs |>
  select(record_name, list_item_count, joined_all_time_count) |>
  arrange(desc(joined_all_time_count))
```

### Starter packs as a natural experiment

Joining through a pack assigns a newcomer hundreds of follows in one
click — a discontinuity in initial conditions that Part 05's designs can
exploit: compare the retention and activity of accounts that joined
through different packs.

## 7. Labels and stackable moderation

Moderation on Bluesky is *stackable*: independent labeler services emit
labels (like `spam` or `graphic-media`) that clients can subscribe to
and act on. The labels themselves are queryable from each labeler.
bskyr has no wrapper for this endpoint, which is a teachable moment in
itself: any XRPC endpoint is just an HTTP URL, so we build the request
directly with httr2 against Bluesky's own moderation service:

```{r labels}
query_labels <- function(patterns, labeler = "https://mod.bsky.app") {
  resp <- httr2::request(paste0(labeler,
                                "/xrpc/com.atproto.label.queryLabels")) |>
    httr2::req_url_query(uriPatterns = patterns, .multi = "explode") |>
    httr2::req_perform() |>
    httr2::resp_body_json()
  map(resp$labels, \(l) tibble(
    uri = l$uri, val = l$val, neg = isTRUE(l$neg)
  )) |> list_rbind()
}

post_labels <- query_labels(head(unique(posts_df$uri), 300))
nrow(post_labels)
```

Zero labels on feed-sampled posts is itself a finding: by the time a
post reaches a curated feed it has survived several moderation filters.
Sampling authors' whole *histories* tells a different story:

```{r history-labels}
set.seed(42)
sampled_dids <- sample(unique(posts_df$author_did),
                       min(50, n_distinct(posts_df$author_did)))

history_labels <- query_labels(paste0("at://", sampled_dids, "/*"))

if (nrow(history_labels) > 0) {
  history_labels |> filter(!neg) |> count(val, sort = TRUE)
} else {
  "No labels found in this sample."
}
```

## 8. Reproducibility: working from the saved archive

```{r reload}
archive <- readRDS("data/feed_posts.rds")

archive |>
  group_by(feed) |>
  summarise(median_likes = median(like_count),
            median_reposts = median(repost_count))
```

## 9. Research design: algorithm audits without platform cooperation

The audit tradition (Sandvig et al. 2014) had to scrape or run sock
puppets because platforms hid their rankers; Bluesky's architecture
makes the *outputs* of every public feed generator directly observable.
What remains unobservable is *personalization*: feeds that rank
relative to a logged-in user can only be audited from that user's
standpoint — exactly the layer Huszár et al. (2022) needed internal
Twitter access to study. A serious audit samples feeds repeatedly over
time, archives everything with retrieval timestamps, and reports the
sampling design alongside the findings.

## Ethics and limitations

Feed samples are snapshots: the same feed an hour later serves different
posts, so archive retrieval times and treat comparisons as
time-specific. Posts surfaced by feeds were written by people who never
chose to be in your sample — the usual minimization rules apply. Label
data is moderation data: publishing which accounts carry which labels
can amplify harm, so aggregate. And remember the deepest limitation:
without logged-in vantage points, personalized curation is invisible.

## Exercises

1. Add a topical feed you use to `CANDIDATE_FEEDS` and rerun the audit.
   Where does it fall on concentration and recency?
2. Sample the Discover feed three times, an hour apart, and measure the
   Jaccard overlap of its post sets — how fast does the feed churn?
3. Compute author-level Gini coefficients per feed as an alternative
   concentration measure.
4. Unpack the most-joined starter pack's list members and check how many
   also appear in the feed samples.
5. Query labels for the *authors* (not posts) in the baseline list feed
   and compare label prevalence with the discovery feeds' authors.

## Research project ideas

* **A longitudinal feed observatory.** Sample a fixed set of feeds daily
  for a month, archiving composition metrics; characterize how editorial
  regimes drift and respond to news events.
* **Starter-pack cohorts.** Identify accounts that joined via packs
  (join counts are public), build matched cohorts (Part 05), and
  estimate the causal effect of initial follow endowments on retention.

## References

### Scholarship

* Sandvig, C., Hamilton, K., Karahalios, K., & Langbort, C. (2014). Auditing algorithms: Research methods for detecting discrimination on internet platforms. *Data and Discrimination*, ICA preconference.
* Huszár, F., Ktena, S. I., O'Brien, C., Belli, L., Schlaikjer, A., & Hardt, M. (2022). Algorithmic amplification of politics on Twitter. *PNAS*, 119(1).
* Bandy, J., & Diakopoulos, N. (2021). More accounts, fewer links: How algorithmic curation impacts media exposure in Twitter timelines. *CSCW*.
* Kleppmann, M., et al. (2024). Bluesky and the AT Protocol: Usable decentralized social media. *arXiv:2402.03239*.

### Documentation

* Feed generators: <https://docs.bsky.app/docs/starter-templates/custom-feeds>
* `com.atproto.label.queryLabels`: <https://docs.bsky.app/docs/api/com-atproto-label-query-labels>
* Stackable moderation: <https://bsky.social/about/blog/03-12-2024-stackable-moderation>
* bskyr: <https://christophertkenny.com/bskyr/>
