---
title: "Bluesky Data Science — Part 00: Introduction"
subtitle: "An R counterpart to the Python notebook, using bskyr"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

This document is the R Markdown counterpart of the Python notebook
[Part 00 - Introduction](../Notebooks/Part%2000%20-%20Introduction.ipynb).
It performs the same first steps — authenticating, retrieving profiles,
feeds, follows, and followers, and (carefully) writing content — using
[bskyr](https://christophertkenny.com/bskyr/), an R interface to the
Bluesky API by Christopher Kenny, together with the
[tidyverse](https://www.tidyverse.org/).

## Background

Researchers studying social media spent a decade building methods on
Twitter's APIs; when that access was shut off, attention turned to open
platforms. Bluesky runs on the [AT Protocol](https://atproto.com), an
open protocol whose data is public by design: profiles, posts, follows,
likes, and reposts can all be retrieved through documented endpoints at
[docs.bsky.app](https://docs.bsky.app). The Python series accesses those
endpoints through the `atproto` SDK; here we use bskyr, which wraps the
same endpoints in R functions that return
[tibbles](https://tibble.tidyverse.org/) ready for tidyverse analysis.

### Code

Install the packages once:

```{r install, eval=FALSE}
install.packages(c("bskyr", "tidyverse", "jsonlite"))
```

Then load them. Each plays a distinct role that you will see in every
part of this series: **bskyr** talks to the Bluesky API, the
**tidyverse** (dplyr, ggplot2, purrr, and friends) manipulates and
plots the data that comes back, and **jsonlite** reads and writes JSON,
the format APIs speak:

```{r libraries}
library(bskyr)
library(tidyverse)
library(jsonlite)
```

### Setup

bskyr authenticates with your handle and an **app password** — create one
at [bsky.app/settings/app-passwords](https://bsky.app/settings/app-passwords),
never your main account password. There are two ways to supply credentials:

1. Set `BLUESKY_APP_USER` and `BLUESKY_APP_PASS` in your `~/.Renviron`
   (bskyr's native mechanism, via `bs_set_user()` and `bs_set_pass()`).
2. Save an `atproto.json` file — the same file the Python notebooks use —
   in the repository root: `{"handle": "you.bsky.social", "password": "..."}`.
   It is gitignored, so credentials never end up in version control.

The shared helper `bsky_client.R` checks both locations and records
whether we are authenticated. **Unlike the Python series, which falls
back to Bluesky's public AppView for read endpoints, bskyr (as of 0.4.x)
sends every request through the authenticated PDS** — so without
credentials the API chunks in this document are skipped when knitting.

## Client

```{r connect}
source("bsky_client.R")
# Without credentials, skip every chunk from here on but keep the prose
knitr::opts_chunk$set(eval = bsky_authed)
```

When credentials are present, `bs_auth()` exchanges them for a session
token that every subsequent bskyr call reuses (it is cached for a few
minutes, so repeated calls don't re-authenticate).

## Profile

`bs_get_profile()` wraps the API endpoint `app.bsky.actor.getProfiles`
and returns one row per requested account. Let's look at our own profile
first. Two pieces of R syntax to get comfortable with right away: the
pipe `|>` sends the result of one step into the next (read it as
"then"), and `glimpse()` prints a wide tibble sideways so every column
fits on screen:

```{r own-profile}
me <- bs_get_profile(bs_get_user())

me |>
  select(handle, display_name, followers_count, follows_count,
         posts_count, created_at) |>
  glimpse()
```

The numbers you see are live: run this tomorrow and the follower count
may differ. The `description`, `avatar`, and other fields are columns
too — a tibble row plays the role the typed `ProfileView` model plays in
the Python series.

### Retrieving other users' profiles

Any public account can be looked up by handle. `bsky.app` is the
platform's official account:

```{r bsky-profile}
bsky_app <- bs_get_profile("bsky.app")

bsky_app |>
  select(handle, display_name, followers_count, posts_count, created_at)
```

`bs_get_profile()` is vectorized: pass several handles at once and each
becomes a row, which is how Part 01 hydrates accounts in batches.

## Posts

`bs_get_author_feed()` wraps `app.bsky.feed.getAuthorFeed` and returns an
account's recent posts, newest first. Each row is one feed item, and the
post's engagement counters (`like_count`, `repost_count`, ...) are
ordinary columns. The author's actual words live somewhere new: a
**list-column** named `record`, where each cell holds the structured
record the author's client wrote (text, timestamp, languages, facets).
List-columns are how tibbles store nested API data, and
`purrr::map_chr()` is how we reach into them — "for each element of
`record`, pull out `text` as a character string":

```{r author-feed}
profile_feed <- bs_get_author_feed("nytimes.com", limit = 100)

nrow(profile_feed)

profile_feed |>
  mutate(text = map_chr(record, "text", .default = NA)) |>
  select(created_at = indexed_at, text, like_count, repost_count) |>
  head(5)
```

This `mutate()` + `map_chr()` pattern appears in every part of the
series — it is the single most useful idiom for working with bskyr's
output.

## Timeline

The home timeline — the reverse-chronological feed of accounts *you*
follow — is defined relative to the logged-in account, which is why this
endpoint always requires authentication:

```{r timeline}
timeline <- bs_get_timeline(limit = 25)

timeline |>
  mutate(text = map_chr(record, "text", .default = NA)) |>
  select(author_handle, text) |>
  head(5)
```

## Relationships

### Follows

`bs_get_follows()` wraps `app.bsky.graph.getFollows`: the accounts an
actor follows, one profile per row.

```{r follows}
follows <- bs_get_follows("nytimes.com", limit = 100)

nrow(follows)
follows |> select(did, handle, display_name) |> head(5)
```

These endpoints are *paginated*: each request returns at most 100
records plus a `cursor` — an opaque bookmark marking where the next page
starts. bskyr tucks the cursor away as an *attribute* of the tibble
(metadata attached to the object, read with `attr()`), and —
conveniently — when you ask for `limit > 100` it follows the cursors for
you, making one request per hundred records:

```{r follows-cursor}
attr(follows, "cursor")$cursor

all_follows <- bs_get_follows("nytimes.com", limit = 1000)
nrow(all_follows)
```

If `nrow(all_follows)` came back smaller than 1,000, that is not an
error — it means we reached the end of the list before reaching our
limit. The limit is a *budget*, not a promise.

Part 01 unpacks exactly what is happening under the hood here.

### Followers

`bs_get_followers()` is the mirror image: accounts that follow the actor.

```{r followers}
followers <- bs_get_followers("nytimes.com", limit = 100)

followers |> select(handle, display_name) |> head(5)
```

## Posting content

Reading is safe to do freely; **writing changes the public record**, so
the chunks below are shown but never executed when this document is knit
(`eval=FALSE`). Run them deliberately, from your own account, when you
mean to.

### Writing a post

```{r post, eval=FALSE}
bs_post(text = "I am posting this from an R Markdown notebook!")
```

### Liking a post

`bs_like()` needs the post's `uri` (its AT-protocol address) and `cid`
(its content hash) — both are columns in any feed tibble:

```{r like, eval=FALSE}
bs_like(
  uri = "at://did:plc:lse7quaysss2d3xxm76mouhd/app.bsky.feed.post/3lareilifis2h",
  cid = "bafyreifacygwvqtrqc7dts2fk67iyv2sa2u3irqdlry7trusq424zmqu7a"
)
```

### Writing a post with an image

`bs_post()` uploads images for you and attaches alt text — which Part 07
shows is both an accessibility practice and a research object:

```{r post-image, eval=FALSE}
bs_post(
  text = "Now we are sending images from a notebook!",
  images = "small_class.jpg",
  images_alt = "Selfie with class"
)
```

## References

* bskyr documentation: <https://christophertkenny.com/bskyr/>
* Kenny, C. T. (2023). bskyr: Interact with 'Bluesky' Social. R package. <https://CRAN.R-project.org/package=bskyr>
* Bluesky HTTP API reference: <https://docs.bsky.app>
* AT Protocol: <https://atproto.com>
* R Markdown documentation: <https://rmarkdown.rstudio.com/docs/>
* Wickham, H., Çetinkaya-Rundel, M., & Grolemund, G. (2023). *R for Data Science* (2nd ed.). <https://r4ds.hadley.nz/>
