---
title: "Part 01 — Collecting Data at Scale: Pagination, Rate Limits, and Research Ethics"
subtitle: "Bluesky Data Science in R"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

This document is the R counterpart of the Python notebook
[Part 01 - Collecting Data at Scale](../../Notebooks/Part%2001%20-%20Collecting%20Data%20at%20Scale/Part%2001%20-%20Collecting%20Data%20at%20Scale.ipynb).
It builds the same documented archive of one account's posts, follows,
and followers — and teaches the same craft of collecting it responsibly
— using [bskyr](https://christophertkenny.com/bskyr/) and the tidyverse.

## Background

Part 00 made single API calls and looked at the first page of results.
Real research needs collections that are *complete where completeness is
feasible and deliberately bounded where it is not*: every account our
case study follows, but only a defensible sample of its 1.3 million
followers. Getting from one page to an archive raises the questions this
part answers: how endpoints paginate, how fast we may call them, how to
flatten nested API responses into tidy data frames, and how to store an
archive so that our future selves — and other researchers — can use it.

### Learning objectives

* Define a sampling frame before collecting anything.
* Explain cursor-based pagination, and use bskyr's automatic paging.
* Wrap API calls in retry logic with exponential backoff.
* Flatten nested responses into tidy tibbles.
* Persist archives to disk with collection metadata for reproducibility.

## Setup

Alongside the core trio from Part 00, this part adds **lubridate**, the
tidyverse's package for parsing and manipulating dates and times — API
timestamps arrive as ISO-8601 strings, and lubridate turns them into
objects we can sort, bin, and plot:

```{r libraries}
library(bskyr)
library(tidyverse)
library(jsonlite)
library(lubridate)
```

As in every part of this series, the shared helper `bsky_client.R` looks
for credentials (an `atproto.json` file or bskyr's environment
variables), authenticates if it finds them, and sets `bsky_authed`. The
`knitr::opts_chunk$set(eval = bsky_authed)` line then makes that flag the
default for the whole document: with credentials every chunk runs;
without them the document still knits, but the API chunks are skipped.

```{r connect}
source("bsky_client.R")
knitr::opts_chunk$set(eval = bsky_authed)
```

## 1. Before you collect: define a sampling frame

A *sampling frame* is the explicit list of units your data could possibly
contain. Ours is one account's complete public activity. Checking the
profile first tells us how much work is ahead:

```{r frame}
HANDLE <- "nytimes.com"

profile <- bs_get_profile(HANDLE)

profile |>
  select(handle, did, created_at, posts_count, follows_count, followers_count)
```

Each paginated endpoint returns at most 100 records per request, so the
counters above translate directly into a request budget. The
`pivot_longer()` call below reshapes the three counter columns into
rows — one row per collection — so we can compute the budget for all
three with a single `mutate()`:

```{r budget}
PAGE_SIZE <- 100

profile |>
  select(posts_count, follows_count, followers_count) |>
  pivot_longer(everything(), names_to = "collection", values_to = "records") |>
  mutate(requests_needed = ceiling(records / PAGE_SIZE))
```

This arithmetic is the lesson. The follows list is two requests. The
posts are a few hundred. The followers of a major newsroom are over a
million records — more than *ten thousand* requests: hours of collection
that would strain the API and hand us a giant file of third parties we
have no question about. Always do this arithmetic *before* you start
collecting, and let it talk you out of "collect everything."

## 2. The anatomy of a paginated endpoint

Ask for one page of the author feed and bskyr returns a tibble of (up to)
100 posts, with the *cursor* — an opaque bookmark marking where the next
page begins — stored as an attribute:

```{r page1}
page1 <- bs_get_author_feed(HANDLE, limit = 100)

nrow(page1)
attr(page1, "cursor")$cursor
```

Passing that cursor back retrieves the next page; the two pages should
not overlap:

```{r page2}
page2 <- bs_get_author_feed(HANDLE, limit = 100,
                            cursor = attr(page1, "cursor")$cursor)

length(intersect(page1$uri, page2$uri))
```

Zero overlap: page 2 picks up exactly where page 1 left off. Walking
cursors until the server stops returning one is all "collecting at
scale" means.

## 3. A reusable collection loop

In the Python series we wrote `fetch_all_pages()` by hand. bskyr builds
the same loop into every paginated function: ask for `limit > 100` and it
issues ceiling(limit/100) requests behind the scenes, following cursors
until the limit — or the end of the data — is reached.

```{r collect-all}
posts_raw <- bs_get_author_feed(HANDLE, limit = 10000)
follows   <- bs_get_follows(HANDLE, limit = 500)
followers <- bs_get_followers(HANDLE, limit = 2000)

c(posts = nrow(posts_raw), follows = nrow(follows), followers = nrow(followers))
```

Note what the caps did. The follows list (a couple hundred accounts)
came back *complete*. The posts came back capped at 10,000 — the most
recent slice of a longer history. And the followers stopped at 2,000 of
more than a million: a deliberate **sample of the most recent
followers**, because no analysis below needs all of them. A `limit` is a
budget and a design decision at once; never write a loop whose only
stopping condition is "the API ran out of data."

## 4. Rate limits, polite clients, and exponential backoff

Bluesky's PDS publishes its rate limits in `RateLimit-*` response
headers. We can read them with a single raw request:

```{r rate-headers}
resp <- httr2::request("https://bsky.social/xrpc/com.atproto.server.describeServer") |>
  httr2::req_perform()

httr2::resp_headers(resp)[c("ratelimit-limit", "ratelimit-remaining",
                            "ratelimit-reset")]
```

### Handling 429s with exponential backoff

When you do hit the limit the server answers `429 Too Many Requests`. The
standard response is *exponential backoff*: wait, retry, and double the
wait each time, so a struggling server gets breathing room instead of a
stampede. The wrapper below uses R's `tryCatch()` — which runs a chunk of
code and, instead of crashing on an error, hands the error to us as an
object we can inspect and respond to. We try the call; if it fails we
sleep `base_wait^attempt` seconds and try again, up to `max_tries`:

```{r backoff}
call_with_backoff <- function(fun, ..., max_tries = 5, base_wait = 2) {
  for (attempt in seq_len(max_tries)) {
    result <- tryCatch(fun(...), error = function(e) e)
    if (!inherits(result, "error")) {
      return(result)
    }
    if (attempt == max_tries) {
      stop(result)
    }
    wait <- base_wait^attempt
    message("Request failed (", conditionMessage(result),
            "); retrying in ", wait, "s")
    Sys.sleep(wait)
  }
}

bsky_profile <- call_with_backoff(bs_get_profile, "bsky.app")
bsky_profile$followers_count
```

(Under the hood, bskyr's HTTP engine [httr2](https://httr2.r-lib.org/)
can also retry transparently via `req_retry()` — worth knowing when you
write raw XRPC requests, as later parts do.)

### Doing more with fewer requests

The cheapest request is the one you never make. `bs_get_profile()`
accepts up to 25 actors per call, so hydrating accounts in batches cuts
your request count 25-fold:

```{r batch}
sample_dids <- head(follows$did, 25)

batch <- bs_get_profile(sample_dids)

batch |> select(handle, followers_count, posts_count) |> head(3)
```

## 5. Flattening nested models into tidy data frames

A feed item nests several objects: the post, its author, the record the
author wrote, and an optional `reason` marking reposts. bskyr has already
widened most of this into columns; what remains is to pull the fields we
want out of the `record` list-column (the `map_chr()` idiom from Part 00)
and drop reposts of other people's words.

Two defensive moves in this chunk deserve explanation, because you will
reuse them constantly. First, the `reason_type` and `reply_parent_uri`
columns *only exist if at least one post on the page was a repost or a
reply* — so we check with `%in% names(...)` before filtering on them.
Second, every extraction supplies a `.default` so that one malformed
record produces an `NA` instead of an error:

```{r flatten}
has_reason <- "reason_type" %in% names(posts_raw)

posts <- posts_raw |>
  filter(if (has_reason) is.na(reason_type) else TRUE) |>
  mutate(
    text       = map_chr(record, "text", .default = NA),
    created_at = ymd_hms(map_chr(record, "created_at", .default = NA),
                         quiet = TRUE),
    langs      = map_chr(record, \(r) paste(unlist(r$langs), collapse = ",")),
    is_reply   = if ("reply_parent_uri" %in% names(posts_raw))
                   !is.na(reply_parent_uri) else FALSE,
    has_embed  = !map_lgl(embed, is.null)
  ) |>
  select(uri, cid, created_at, text, is_reply,
         like_count, repost_count, reply_count, quote_count,
         langs, has_embed) |>
  arrange(created_at)

glimpse(posts)
```

One row per post, one column per feature, timestamps parsed — the *tidy
data* shape every later part consumes.

## 6. Persisting archives: CSV, RDS, and metadata

Collection is slow and the network changes under you, so collect once and
analyze from disk. CSV travels everywhere; RDS preserves R types exactly
(the Python series uses Parquet for the same reason):

```{r persist}
dir.create("data", showWarnings = FALSE)

write_csv(posts, "data/posts.csv")
saveRDS(posts, "data/posts.rds")
write_csv(select(follows, did, handle, display_name), "data/follows.csv")
write_csv(select(followers, did, handle, display_name), "data/followers.csv")

file.size(c("data/posts.csv", "data/posts.rds")) / 1024
```

### Recording collection metadata

An archive without provenance is a liability. Write down what you
collected, when, how, and as whom:

```{r metadata}
metadata <- list(
  account = list(handle = HANDLE, did = profile$did),
  collected_at_utc = format(now("UTC"), "%Y-%m-%dT%H:%M:%SZ"),
  bskyr_version = as.character(packageVersion("bskyr")),
  collections = list(
    posts = list(endpoint = "app.bsky.feed.getAuthorFeed",
                 n_records = nrow(posts),
                 notes = "Original posts only; reposts excluded."),
    follows = list(endpoint = "app.bsky.graph.getFollows",
                   n_records = nrow(follows)),
    followers = list(endpoint = "app.bsky.graph.getFollowers",
                     n_records = nrow(followers))
  )
)

write_json(metadata, "data/collection_metadata.json",
           pretty = TRUE, auto_unbox = TRUE)
```

### Proving the round trip

```{r roundtrip}
reloaded <- readRDS("data/posts.rds")

identical(reloaded, posts)
```

## 7. A first look at the archive

The payoff: a complete posting history, summarized in a few lines of
dplyr and ggplot2. `floor_date()` rounds each timestamp down to the
first of its month, and `count()` tallies posts per month — the
month-binning pattern that Part 05 builds a whole time-series toolkit
on:

```{r posts-per-month, fig.width=10, fig.height=4}
posts |>
  count(month = floor_date(created_at, "month")) |>
  ggplot(aes(month, n)) +
  geom_col(fill = "steelblue") +
  labs(x = "Month", y = "Posts",
       title = paste0("Posts per month by @", HANDLE)) +
  theme_minimal()
```

```{r engagement-summary}
posts |>
  summarise(
    share_replies = mean(is_reply),
    share_embeds  = mean(has_embed),
    median_likes  = median(like_count, na.rm = TRUE),
    max_likes     = max(like_count, na.rm = TRUE)
  )
```

And the relationship between whom the account follows and who follows
it. `intersect()` finds the DIDs present in both lists — accounts with a
*mutual* (reciprocated) tie, a concept Part 03 returns to as
reciprocity. One honesty note: our followers table is a 2,000-account
sample of more than a million, so `share_followed` describes that recent
sample, not the full follower population:

```{r mutuals}
mutuals <- intersect(follows$did, followers$did)

tibble(
  follows = nrow(follows),
  followers = nrow(followers),
  mutuals = length(mutuals),
  share_follows_back = length(mutuals) / nrow(follows),
  share_followed     = length(mutuals) / nrow(followers)
)
```

## Ethics and limitations

Everything we collected is public, but **public is not the same as
consented**: the people in these archives did not agree to be studied
(boyd & Crawford 2012; Fiesler & Proferes 2018). Practice *data
minimization* — collect only what your question needs, prefer aggregates,
and think hard before storing third parties (the followers list is a
list of people who never interacted with you). People delete posts and
accounts; an archive freezes what the network has moved past, so decide
in advance how you will honor deletion and withdrawal. Your collection
also runs under Bluesky's terms of service and, if you are doing human
subjects research, your IRB's rules. Methodologically, remember that one
account is a case study, the author feed omits deleted and moderated
posts, and engagement counters are point-in-time snapshots that change
after collection.

## Exercises

1. Re-run the collection for a different public account (e.g.
   `bsky.app`) and compare its posting cadence to this one.
2. Modify `call_with_backoff()` to give up immediately on errors that
   will never succeed on retry (e.g. an account that does not exist).
3. Use the `subject_*` columns of `bs_get_follows()` to check whether the
   follows list itself records when each follow was created.
4. Compute the share of posts that are replies, by month — is the account
   becoming more or less conversational?
5. Compare the on-disk size of the posts archive as CSV, RDS, and (with
   the [arrow](https://arrow.apache.org/docs/r/) package) Parquet.

## Research project ideas

* **Activity and audience growth.** Collect monthly posting volume and
  follower counts for a panel of public accounts and model how posting
  consistency relates to audience growth.
* **Reciprocity norms.** For accounts of different sizes, measure the
  share of follows that are mutual; test whether reciprocity declines
  with audience size, as it did on Twitter.

## References

### Scholarship

* boyd, d., & Crawford, K. (2012). Critical questions for big data. *Information, Communication & Society*, 15(5), 662–679.
* Fiesler, C., & Proferes, N. (2018). "Participant" perceptions of Twitter research ethics. *Social Media + Society*, 4(1).
* Freelon, D. (2018). Computational research in the post-API age. *Political Communication*, 35(4), 665–668.
* Zimmer, M. (2010). "But the data is already public": On the ethics of research in Facebook. *Ethics and Information Technology*, 12(4), 313–325.

### Documentation

* bskyr: <https://christophertkenny.com/bskyr/>
* `app.bsky.feed.getAuthorFeed`: <https://docs.bsky.app/docs/api/app-bsky-feed-get-author-feed>
* Bluesky rate limits: <https://docs.bsky.app/docs/advanced-guides/rate-limits>
* httr2 retries: <https://httr2.r-lib.org/reference/req_retry.html>
* R Markdown: <https://rmarkdown.rstudio.com/docs/>
