---
title: "Part 04 — Threads, Engagement, and Information Diffusion"
subtitle: "Bluesky Data Science in R"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

This document is the R counterpart of the Python notebook
[Part 04 - Threads Engagement and Diffusion](../../Notebooks/Part%2004%20-%20Threads%20Engagement%20and%20Diffusion/Part%2004%20-%20Threads%20Engagement%20and%20Diffusion.ipynb).
It reconstructs a reply *cascade* from `app.bsky.feed.getPostThread`,
measures its shape with [igraph](https://r.igraph.org/), traces who
liked, reposted, and quoted the post and when, and shows that engagement
across posts is radically unequal.

### Learning objectives

* Traverse nested thread views recursively into a reply tree.
* Compute cascade metrics: size, depth, breadth, structural virality.
* Analyze the timing of engagement: the burst and the decay.
* Plot heavy-tailed engagement distributions correctly (CCDFs, log-log).
* Recognize selection on the dependent variable in diffusion research.

## Setup

```{r libraries}
library(bskyr)
library(tidyverse)
library(jsonlite)
library(lubridate)
library(igraph)
```

As in every part of this series, the shared helper `bsky_client.R` looks
for credentials (an `atproto.json` file or bskyr's environment
variables), authenticates if it finds them, and sets `bsky_authed`. The
`knitr::opts_chunk$set(eval = bsky_authed)` line then makes that flag the
default for the whole document: with credentials every chunk runs;
without them the document still knits, but the API chunks are skipped.

```{r connect}
source("bsky_client.R")
knitr::opts_chunk$set(eval = bsky_authed)

COLLECTED_AT <- format(now("UTC"), "%Y-%m-%dT%H:%M:%SZ")
```

### A note on parsing, and a platform in flux

The Python notebook opens with a lesson about *strict* SDK parsing:
typed models crash when Bluesky ships an embed type the SDK hasn't met.
R sidesteps the issue by temperament — `clean = FALSE` hands us plain
nested lists, and lists don't validate anything. The price is that *we*
must know the structure; the reward is that nothing the platform adds
can crash us. For deeply nested responses like threads, this part works
with raw lists and `purrr`.

## 1. Finding a study post

We need a post with a deep conversation under it. The official
`bsky.app` account reliably draws large reply threads:

```{r find-post}
feed <- bs_get_author_feed("bsky.app", limit = 100)

is_reply  <- if ("reply_parent_uri" %in% names(feed))
               !is.na(feed$reply_parent_uri) else FALSE
is_repost <- if ("reason_type" %in% names(feed))
               !is.na(feed$reason_type) else FALSE

originals <- feed |>
  filter(!is_reply, !is_repost) |>
  mutate(text = map_chr(record, "text", .default = ""))

originals |>
  select(reply_count, like_count, repost_count, quote_count) |>
  summary()

case_study <- originals |> slice_max(reply_count, n = 1, with_ties = FALSE)

case_study |> select(uri, reply_count, like_count, repost_count, text)
```

## 2. From thread views to reply trees

`bs_get_post_thread()` returns the post and its replies nested `depth`
levels down. With `clean = FALSE` the response is a plain list whose
`$thread$replies` elements each contain their own `$replies` — a
recursive structure, which calls for a *recursive function*: one that
handles a single node (build its row) and then calls itself on each of
that node's children. Two idioms to note: `%||%` means "use the left
side unless it is NULL, then use the right side" (our defense against
missing fields), and a node without a `$post` is the stub left behind by
a deleted or blocked reply — we record nothing for it, but its existence
is a data-quality fact worth remembering:

```{r thread-to-edges}
thread_to_edges <- function(node, parent_uri = NA_character_) {
  post <- node$post
  if (is.null(post)) {
    # a notFoundPost or blockedPost stub: the hole a deletion leaves
    return(tibble(uri = character(0)))
  }
  row <- tibble(
    uri = post$uri,
    parent_uri = parent_uri,
    author_did = post$author$did,
    created_at = post$record$createdAt %||% NA_character_,
    reply_count = post$replyCount %||% 0L,
    like_count = post$likeCount %||% 0L,
    truncated = (post$replyCount %||% 0L) > 0 && length(node$replies) == 0
  )
  children <- map(node$replies, thread_to_edges, parent_uri = post$uri)
  bind_rows(row, children)
}

response <- bs_get_post_thread(case_study$uri, depth = 6, clean = FALSE)
tree <- thread_to_edges(response[[1]]$thread)

c(posts = nrow(tree),
  truncated_branches = sum(tree$truncated))
```

A single request only reaches `depth` levels, so deep conversations get
*truncated*: nodes that claim replies we have not seen. The fetcher
below keeps a **frontier** — a queue of truncated nodes still to expand
— and works through it with follow-up requests until the frontier is
empty or the request budget runs out. The `seen` vector prevents
double-counting posts that appear in overlapping responses:

```{r fetch-full-tree}
fetch_thread_tree <- function(uri, depth = 6, max_requests = 40) {
  requests <- 0
  frontier <- uri
  seen <- character(0)
  pieces <- list()
  while (length(frontier) > 0 && requests < max_requests) {
    target <- frontier[[1]]
    frontier <- frontier[-1]
    response <- tryCatch(
      bs_get_post_thread(target, depth = depth, clean = FALSE),
      error = function(e) NULL
    )
    requests <- requests + 1
    Sys.sleep(0.1)
    if (is.null(response)) next
    piece <- thread_to_edges(response[[1]]$thread)
    piece <- filter(piece, !uri %in% seen)
    seen <- c(seen, piece$uri)
    pieces <- c(pieces, list(piece))
    frontier <- c(frontier, piece$uri[piece$truncated])
  }
  out <- bind_rows(pieces)
  attr(out, "requests_made") <- requests
  attr(out, "frontier_remaining") <- length(frontier)
  out
}

tree <- fetch_thread_tree(case_study$uri, depth = 6, max_requests = 40)

c(posts = nrow(tree),
  requests = attr(tree, "requests_made"),
  unexpanded = attr(tree, "frontier_remaining"))
```

### How much of the cascade did we get?

Every node advertises its `reply_count`; comparing promised replies to
retrieved edges is an honest coverage check:

```{r coverage}
promised <- sum(tree$reply_count)
retrieved <- sum(!is.na(tree$parent_uri))

c(promised = promised, retrieved = retrieved,
  coverage = round(retrieved / promised, 3))
```

## 3. Cascade metrics

Build the tree as an igraph object and measure its shape. **Structural
virality** (Goel et al. 2016) is the mean shortest-path distance between
all pairs of nodes: low for broadcast (everyone replies to the root),
high for chains that spread hand-to-hand.

```{r metrics}
edges <- tree |> filter(!is.na(parent_uri)) |> select(parent_uri, uri)
Gt <- graph_from_data_frame(edges, directed = TRUE,
                            vertices = tibble(name = tree$uri))

depths <- distances(Gt, v = case_study$uri, mode = "out")[1, ]

structural_virality <- function(g) {
  mean_distance(as_undirected(g), unconnected = FALSE)
}

tibble(
  size = vcount(Gt),
  depth = max(depths[is.finite(depths)]),
  max_breadth = max(table(depths[is.finite(depths)])),
  unique_repliers = n_distinct(tree$author_did),
  structural_virality = structural_virality(Gt)
)
```

Calibrate the structural-virality number against synthetic extremes of
the same size:

```{r calibrate}
n <- vcount(Gt)
c(
  star_broadcast = structural_virality(make_star(n, mode = "out")),
  observed = structural_virality(Gt),
  chain_viral = structural_virality(make_ring(n, circular = FALSE))
)
```

```{r depth-profile, fig.width=8, fig.height=4}
tibble(depth = depths[is.finite(depths)]) |>
  count(depth) |>
  ggplot(aes(depth, n)) +
  geom_col(fill = "steelblue") +
  labs(x = "Depth (replies away from the root)", y = "Posts at this depth",
       title = "Depth profile of the case-study reply tree") +
  theme_minimal()
```

### Drawing the reply tree

```{r draw-tree, fig.width=9, fig.height=9}
set.seed(42)
layout <- layout_as_tree(Gt, root = case_study$uri, circular = TRUE)

plot(Gt, layout = layout,
     vertex.size = ifelse(V(Gt)$name == case_study$uri, 6, 1.5),
     vertex.color = ifelse(V(Gt)$name == case_study$uri,
                           "crimson", "steelblue"),
     vertex.label = NA, edge.arrow.size = 0,
     edge.color = adjustcolor("gray", alpha.f = 0.4),
     main = paste0("Reply tree of the case-study post (",
                   vcount(Gt), " posts retrieved)"))
```

## 4. Who engages, and when

Replying is one engagement among several. Likes carry two timestamps —
`createdAt` from the liker's client and `indexedAt` from the network —
which lets us reconstruct the post's attention curve. With
`clean = FALSE`, `bs_get_post_likes()` hands back a list of raw response
pages; each page has a `$likes` list whose elements we flatten into one
row apiece, then convert "when was this like indexed" into "hours since
the post was born":

```{r likes}
likes_raw <- bs_get_post_likes(case_study$uri, limit = 3000, clean = FALSE)

likes <- map(likes_raw, \(page) {
  map(page$likes, \(l) tibble(
    did = l$actor$did,
    handle = l$actor$handle,
    created_at = l$createdAt,
    indexed_at = l$indexedAt
  )) |> list_rbind()
}) |> list_rbind()

post_time <- ymd_hms(case_study$record[[1]]$created_at, quiet = TRUE)

likes <- likes |>
  mutate(
    indexed_at = ymd_hms(indexed_at, quiet = TRUE),
    hours_since_post = as.numeric(difftime(indexed_at, post_time,
                                           units = "hours"))
  ) |>
  filter(hours_since_post >= 0) |>
  arrange(hours_since_post)

c(likes_retrieved = nrow(likes),
  reported_at_collection = case_study$like_count)
```

### The burst, then the decay

```{r burst, fig.width=8, fig.height=4}
likes |>
  mutate(cumulative = row_number()) |>
  ggplot(aes(hours_since_post / 24, cumulative)) +
  geom_line(color = "steelblue") +
  labs(x = "Days since post", y = "Cumulative likes",
       title = "Cumulative likes over time") +
  theme_minimal()

likes |>
  filter(hours_since_post <= 72) |>
  ggplot(aes(hours_since_post)) +
  geom_histogram(bins = 72, fill = "steelblue") +
  labs(x = "Hours since post (first 72)", y = "Likes per hour",
       title = "The initial burst") +
  theme_minimal()
```

Most attention arrives within hours — the universal "burst then decay"
shape of social media attention.

### Reposters and quoters

```{r reposts-quotes}
reposters <- bs_get_reposts(case_study$uri, limit = 500)
quotes <- bs_get_quotes(case_study$uri, limit = 200)

c(reposters = nrow(reposters), quotes = nrow(quotes),
  quote_likes = sum(quotes$like_count, na.rm = TRUE))
```

Each quote post can seed its own cascade — diffusion on Bluesky is a
forest, not a single tree.

### Do the same people like and repost?

```{r overlap}
both <- intersect(likes$did, reposters$did)

c(likers = n_distinct(likes$did),
  reposters = n_distinct(reposters$did),
  both = length(both),
  share_of_reposters_who_liked = round(length(both) / nrow(reposters), 2))
```

## 5. Heavy-tailed engagement across posts

Single cascades mislead; the *distribution* across posts is the durable
fact. Compare the official platform account with a national newsroom, on a
complementary CDF with log-log axes:

```{r heavy-tails, fig.width=10, fig.height=4.5}
nyt_feed <- bs_get_author_feed("nytimes.com", limit = 200)
nyt_originals <- nyt_feed |>
  filter(if ("reason_type" %in% names(nyt_feed))
           is.na(reason_type) else TRUE)

ccdf <- function(x) {
  x <- sort(x[!is.na(x)])
  tibble(value = x, p = 1 - (seq_along(x) - 1) / length(x))
}

bind_rows(
  ccdf(originals$like_count) |> mutate(account = "bsky.app"),
  ccdf(nyt_originals$like_count) |> mutate(account = "nytimes.com")
) |>
  filter(value > 0) |>
  ggplot(aes(value, p, color = account)) +
  geom_step() +
  scale_x_log10() + scale_y_log10() +
  labs(x = "Likes per post", y = "P(X ≥ x)",
       title = "Like-count CCDF (log-log)") +
  theme_minimal()
```

```{r top10-share}
bind_rows(
  mutate(originals, account = "bsky.app"),
  mutate(nyt_originals, account = "nytimes.com")
) |>
  group_by(account) |>
  summarise(
    median_likes = median(like_count, na.rm = TRUE),
    max_likes = max(like_count, na.rm = TRUE),
    top10pct_share = {
      l <- sort(like_count, decreasing = TRUE)
      sum(head(l, ceiling(length(l) * .1))) / max(sum(l), 1)
    }
  )
```

A handful of posts capture most of the attention — the heavy tail that
makes "the average post" a misleading unit of analysis.

## 6. Saving the data

```{r save}
dir.create("data", showWarnings = FALSE)

bind_rows(
  mutate(originals, account = "bsky.app"),
  mutate(nyt_originals, account = "nytimes.com")
) |>
  select(account, uri, reply_count, like_count,
         repost_count, quote_count) |>
  mutate(collected_at = COLLECTED_AT) |>
  write_csv("data/engagement_summary.csv")

likes |>
  transmute(liker_index = row_number(),   # anonymous index, not an identity
            hours_since_post, collected_at = COLLECTED_AT) |>
  write_csv("data/case_study_likes.csv")

fs::dir_info("data")[, c("path", "size")]
```

## 7. Research design: studying diffusion without sampling on success

The fatal error in diffusion research is **selecting on the dependent
variable**: studying only the cascades that went viral and "discovering"
the properties of viral content. Goel et al. (2012) showed that when you
sample *all* cascades, the overwhelming majority are tiny — depth one,
size one. Vosoughi, Roy, & Aral (2018) compared true and false news by
conditioning on *content verified externally*, not on success. The
honest design defines the population of posts first (an account's full
history, a time window from the firehose — Part 10) and treats cascade
size as the outcome, never the sampling criterion.

## Ethics and limitations

Reply trees include ordinary people who answered a prominent account;
analyze the *structure*, and avoid republishing individual repliers'
words or identities (our saved likes table keeps only timing, not
identities). Coverage is never complete: deleted and blocked replies
leave holes, the thread endpoint truncates, and counters drift. The
case-study post was selected *because* it was successful — fine for
illustrating methods, fatal for causal claims, as Section 7 explains.

## Exercises

1. Re-run the cascade collection for the *second*-most-replied post. How
   do its depth and structural virality compare?
2. Compute the share of all replies that are depth-1. What does it say
   about broadcast versus conversation?
3. Plot likes-per-hour for the first week on log-log axes and estimate
   the decay exponent with a linear fit.
4. Use `bs_get_quotes()` to fetch the largest quote post's own thread.
   How big is the secondary cascade?
5. Build the CCDF comparison with `repost_count` instead of likes. Is
   reposting more or less unequal than liking?

## Research project ideas

* **Structural virality on Bluesky.** Sample posts from the firehose
  (Part 10), reconstruct their cascades after one week, and replicate
  Goel et al.'s (2016) distribution of structural virality on an open
  platform.
* **Quote-post framing.** Compare the language of quote posts to the
  posts they quote (Part 06's methods) to measure how reframing changes
  as information diffuses.

## References

### Scholarship

* Goel, S., Watts, D. J., & Goldstein, D. G. (2012). The structure of online diffusion networks. *EC '12*.
* Goel, S., Anderson, A., Hofman, J., & Watts, D. J. (2016). The structural virality of online diffusion. *Management Science*, 62(1), 180–196.
* Vosoughi, S., Roy, D., & Aral, S. (2018). The spread of true and false news online. *Science*, 359(6380), 1146–1151.
* Crane, R., & Sornette, D. (2008). Robust dynamic classes revealed by measuring the response function of a social system. *PNAS*, 105(41).

### Documentation

* `app.bsky.feed.getPostThread`: <https://docs.bsky.app/docs/api/app-bsky-feed-get-post-thread>
* `app.bsky.feed.getLikes`: <https://docs.bsky.app/docs/api/app-bsky-feed-get-likes>
* igraph: <https://r.igraph.org/>
* bskyr: <https://christophertkenny.com/bskyr/>
